---
title: "Computing for the Social Sciences: Lecture 4"
subtitle: "Topics: Graphs for Data Analysis"
author: "Sabrina Nardin, Fall 2025"
format:
  revealjs:
    theme: simple
    slide-number: true
    incremental: false
    css: ../style/styles.css
    code-overflow: wrap
    highlight-style: github
    ratio: 16:9
    chalkboard: true
    toc: false
    center: true
---


# Agenda {.center}

::: {.agenda-list}
1. The `scorecard` Dataset  
2. Match Graph Types to Variable Types  
3. In-class Practice  
:::

<!--
```{r, echo=FALSE, results='asis'}
cat('<span style="font-size: 0.8em; color: #666;">These slides were last updated on', format(Sys.Date(), '%B %d, %Y'), '</span>')
```
-->

---

## Our Goal Today: Using Graphs for Exploratory Data Analysis

- **Display variation and co-variation:** Learn how to visualize the distribution of a single variable (`cost` to attend a school) and of two or more variables (`cost` and `admrate`).

- **Match graph type to variable type:** Choose appropriate graph types depending on whether variables are categorical (school `type`), continuous (`cost`), or both.

- **Interpret the graph:** Practice describing what a graph reveals, such as trends, group differences, and outliers.


# 1. The scorecard Dataset {.slide .center .middle}

---

## About the scorecard dataset

The U.S. Department of Education collects annual statistics on colleges and universities in the United States: <https://collegescorecard.ed.gov/data>

This dataset includes variables such as: 

- `name`: name of the school 
- `state`: state where the school is located  
- `type`: school type (e.g., Public, Private Nonprofit, Private For-Profit)  
- `admrate`: admission rate (e.g., 0.91 = 91%)  
- `cost`: published cost of attendance  
- `netcost`: net cost of attendance after financial aid  
- `satavg`: average SAT score of admitted students  
- `...`: many additional variables

We’ll focus on a subset of this data from the 2018–2019 academic year.

---

## About the scorecard dataset

```r
library(tidyverse)
library(rcis)
data(scorecard)
glimpse(scorecard)
```

```{r}
library(tidyverse)
library(rcis)
data(scorecard)
glimpse(scorecard)
```

---

## The scorecard dataset: making a plot

Given this question **"Which type of college has the highest average SAT score?"**, and the task of answering it by creating a `ggplot2` graph. 

How would you approach this task?

- identify the variable(s) you need
- find them in the data
- identify the variable(s) types (mainly: categorical or continuous)
- use this info to select the most appropriate plot 
- think about the code
- generate the graph and interpret it

---

## Which type of college has the highest average SAT score?

#### Answer using a boxplot

```r
ggplot(data = scorecard, mapping = aes(x = type, y = satavg)) +
  geom_boxplot()
```

```{r}
ggplot(data = scorecard, mapping = aes(x = type, y = satavg)) +
  geom_boxplot()
```

#### [What's a boxplot?](https://datatab.net/tutorial/box-plot){target="_blank"}


---

## Which type of college has the highest average SAT score?

#### Answer using a faceded histogram

```r
ggplot(data = scorecard, mapping = aes(x = satavg)) +
  geom_histogram() + 
  facet_wrap(facets = vars(type))
```

```{r}
ggplot(data = scorecard, mapping = aes(x = satavg)) +
  geom_histogram() + 
  facet_wrap(facets = vars(type))
```

#### [What's a histogram?](https://datatab.net/tutorial/charts){target="_blank"}

---

## Which type of college has the highest average SAT score?

Let's interpret these graphs:

#### What do these graphs reveal about average SAT scores by type of college? Do they reveal the same info? 

---

## Which type of college has the highest average SAT score?

::: {.callout-note title="Interpreting plots is as important as writing code"}

**General interpretation:**

Private nonprofit schools have the highest average SAT scores, closely followed by public schools, and then private for-profit schools.

**More refined interpretation:**

The histogram shows that each school type has a different number of colleges: private for-profit schools are much fewer than the others. This difference in sample size isn’t visible in the boxplot, but it matters: averages from small groups may not reflect the broader category. Seeing the full distribution gives a more complete picture.

:::

---

## Asking More Questions with dplyr (sneak peek for next week!)


How many schools are in **each type?** 
```r
scorecard %>% count(type) 
```

Which schools are categorized as **private for-profit?**
```r
scorecard %>%
  filter(type == "Private, for-profit") %>%
  select(name, state, type, satavg, cost)
```

What about the **University of Chicago?**
```r
scorecard %>% filter(name == "University of Chicago")
```

Etc. 

<!--
The code below creates a summary dataset with `type` of university as rows and the `satavg` mean as a column:

```{r}
scorecard %>% 
  group_by(type) %>%
  summarize(mean_type = mean(satavg, na.rm = TRUE))
```
-->

---

## Once you draw a plot, ask yourself:

::: columns
::: column

#### Substantive questions:
- What does this graph tell?
- Are there patterns? Outliers?
- What hypotheses can I generate?
- What else I want to know?
- Should I dig deeper with `dplyr`?
- Should I wrangle my data and plot again?
- Etc.

:::
::: column

#### Stylistic questions:
- Is the chosen plot appropriate (matches variable type)?
- Is the plot clear and easy to understand?
- Is it too busy or too simple?
- Could it be improved? (titles, labels, colors, etc.)
- Can I make an additional plot with these variables?
- Etc.

:::
:::

# 2. Match Graph Types to Variable Types {.slide .center .middle}

---

## Types of Visualizations and Best Graph Types

### 1. Which kind of variation do I want to represent?

- **Within single variable**
- **Among Two variables**
- **Among Three variables**

<br>

### 2. What type(s) are my variables?

- **Continuous** (e.g., `satavg` average SAT score, `cost` published cost)
- **Categorical** (e.g., `type` school type, `state` U.S. state)
- Other types (often treated as categorical): ordinal, nominal, binary

</br>

*Tip: Ask yourself these two questions before jumping into the code!*

---

## Univariate (One Variable)

**To show how values vary within a single variable**:

- One **continuous** variable → **Histogram**
- One **categorical** variable → **Bar Chart**

<br>

::: {.left}

| Function           | What Gets Counted                                 | When to Use             |
|--------------------|---------------------------------------------------|--------------------------|
| `geom_histogram()` | How many values fall into each numeric bin        | When `x` is continuous   |
| `geom_bar()`       | How many observations are in each category        | When `x` is categorical  |

:::

</br>

---

## Bivariate (Two Variables)

**To show how two variables co-vary**:

- Two **continuous** variables → **Scatterplot**
- One **categorical** + one **continuous** → **Box Plot**
- Two **categorical** variables → **Grouped or Stacked Bar Chart**

---

## Multivariate (Three Variables)

**To compare patterns across subgroups**:

- One **categorical** + two **continuous** → **Faceted Scatterplot**
- Two **categorical** + one **continuous** → **Grouped Box Plot**


---

## Main Graph Types and When to Use Them

| Type of Graph        | `ggplot2` Function                                                              | Input Variables                                                                 | Goal                                  |
|----------------------|----------------------------------------------------------------------------------------|----------------------------------------------------------------------------------|----------------------------------------------------------|
| Histogram            | [geom_histogram()](https://ggplot2.tidyverse.org/reference/geom_histogram.html)       | One **continuous** (e.g., age, income)                              | Show the **distribution** of values                     |
| Bar Chart            | [geom_bar()](https://ggplot2.tidyverse.org/reference/geom_bar.html)                   | One **categorical** (e.g., region, gender)                           | Show **frequencies** or **counts** of categories        |
| Stacked Bar Chart    | [geom_bar() + fill](https://ggplot2.tidyverse.org/reference/geom_bar.html)            | Two **categorical** variables (e.g., region by gender)                         | Compare **parts of a whole** across categories          |
| Scatterplot          | [geom_point()](https://ggplot2.tidyverse.org/reference/geom_point.html)               | Two **continuous** variables (e.g., height vs. weight, price vs. rating)        | Show **relationship** or **correlation**                |
| Box Plot             | [geom_boxplot()](https://ggplot2.tidyverse.org/reference/geom_boxplot.html)           | One **continuous** + one **categorical** (e.g., income by gender)           | Compare **distributions**, spot **outliers**            |
| Faceted Scatterplot  | [facet_wrap() + geom_point()](https://ggplot2.tidyverse.org/reference/facet_wrap.html) | Two **continuous** + one **categorical** (e.g., by country or year)             | Compare **patterns across groups**                      |

*These are the most common combinations. There are more options as you explore deeper!*


# 3. Practice {.slide .center .middle}

---

## 💻 Practice: Make Graphs!

On the next slide, there are a list of graphs to create. 

In small groups, use the scorecard dataset to create the most appropriate graph for each one.

- **Before plotting:** Think about the variable types and the kind of variation you want to represent. (Use the reference slides as a guide.)

- **While plotting:** Keep it simple, as you would for an initial Exploratory Data Analysis (no need to add labels, legends, color, etc.). Look up the documentation for each graph: you do not need to remember all the code!

- **After plotting:** Stare at your graph... look for patterns, outliers, or any notable features, and interpret what graph tells about the data.

---

## 💻 Practice

<!-- For regular CSS class, add some dpyr commands in task and also bring back TASK 4  Display the total number of schools in the U.S. by school type (n = 3) and by state (n = 54). *Note: the initial graph you generate here may lack visual appeal. Focus on identifying potential improvements rather than implementing them for now.* -->

Share your solutions here <https://codeshare.io/5z6PkW>

Graph 1: Plot the annual total cost of school attendance across the U.S. (`cost`)

Graph 2: Plot the total number of schools in the U.S. by school type (`type`)

Graph 3: Plot the annual total cost and net cost of attendance to schools in the U.S. (`cost`, `netcost`)

Graph 4: Plot the annual total cost of attendance by school type (`cost`, `type`)

Graph 5: Plot the annual total cost of attendance and net cost of attendance by school type (`cost`, `netcost`, `type`)

---

## 💻 Download Today Materials

Download today’s class materials from our website for: 

- further insights into these tasks
- additional practice exercises (optional)

---

## Takeaways and Tips

Follow this approach to move from quick exploration to polished, communicative plots:

::: columns
::: column

#### When exploring your data:
- Start with simple and plots to get a sense of the distribution and relationships
- Focus on interpretation first, not style
- Ask questions: What stands out? Are there outliers, trends, or surprises? What else do you want to know?

:::
::: column

#### Once you settle on a plot:
- Refine the code to improve readability and aesthetics
- Add styling elements (labels, scales, legends, and themes. etc. as needed)
- Use polished plots for assignments, reports, or presentations

:::
:::

*Typically, researchers create many exploratory plots and only a few make it to the final report.*

---

## Recap: What We Learned Today

- Pick graph types that match the types of variables in your data
- Read and interpret visualizations
- Pracice using `ggplot` and `dplyr`

---

## To print these slides as pdf

Click on the icon bottom-right corner \> Tools \> PDF Export Mode \> Print as a Pdf

