---
title: "Computing for the Social Sciences: Lecture 4"
subtitle: "Topics: Graphs for Data Analysis with ggplot2"
author: "Sabrina Nardin, Summer 2026"
format:
  revealjs:
    theme: simple
    slide-number: true
    incremental: false
    css: ../style/styles.css
    code-overflow: wrap
    highlight-style: github
    ratio: 16:9
    chalkboard: true
    toc: false
    center: true
---


# Agenda {.center}

::: {.agenda-list}
1. The "scorecard" Dataset  
2. Match Graph Types to Variable Types  
3. In-class Practice  
:::

<span style="font-size: 0.8em; color: #666;">
*Slides last updated on `r format(Sys.Date(), '%B %d, %Y')` (removed repetead slides; minor updates to activity questions). Slides authored by Sabrina Nardin. AI used to polish slides style and fix typos.*
</span>

<!--
### Slides are in progress and will be updated before class.

```{r, echo=FALSE, results='asis'}
cat('<span style="font-size: 0.8em; color: #666;">These slides were last updated on', format(Sys.Date(), '%B %d, %Y'), '</span>')
```

-->

---

## Accessing Poll Everywhere Questions for Participation credits

You have three ways to access Poll Everywhere questions:

1. **Use the Zoom chat link**
   I will post each individual question link in the Zoom chat during class.

2. **Use the main Poll Everywhere link**
   You can also go directly to: **https://pollev.com/nardin**
   This link will not change. Once I activate a specific question, it should appear there automatically.

3. **Use the weekly participation document after class**
   I will add all Poll Everywhere links to the [weekly participation document](https://docs.google.com/document/d/1f5jZ1ynOMV-axF-toNtec-KyOwhLvSYZCx0BlPaslxQ/edit?usp=sharing) (also linked under “Before Class” for each lecture). If you miss something, you can still submit your answers before the next class

On Poll Everywhere: make sure your responses are submitted under your name, not as “Guest.”

---

## Our Goal Today: Graphs for Exploratory Data Analysis

::: {.columns}

::: {.column width="33%"}
### 1. See patterns

Use graphs to explore *variation* and *co-variation* in the data.
:::

<!-- Learn how to visualize the distribution of a single variable (`cost` to attend a school) and of two or more variables (`cost` and `admrate`).
-->

::: {.column width="33%"}
### 2. Choose the graph

Match the plot to the variable type:

- categorical (e.g. school type)
- continuous (e.g., cost of attending a school)
- both
:::

::: {.column width="33%"}
### 3. Explain the graph

Describe what you see:

- trends
- differences
- outliers
:::

:::

# 1. The scorecard Dataset {.slide .center .middle}

---

## About the scorecard dataset

The U.S. Department of Education collects annual statistics on colleges and universities in the United States: <https://collegescorecard.ed.gov/data>

This dataset includes variables such as: 

- `name`: name of the school 
- `state`: state where the school is located  
- `type`: school type (e.g., Public, Private Nonprofit, Private For-Profit)  
- `admrate`: admission rate (e.g., 0.91 = 91%)  
- `cost`: published cost of attendance  
- `netcost`: net cost of attendance after financial aid  
- `satavg`: average SAT score of admitted students  
- `...`: many additional variables

We’ll focus on a subset of this data from the 2018–2019 academic year.

---

## About the scorecard dataset

```r
library(tidyverse)
library(rcis)
data(scorecard)
glimpse(scorecard)
```

```{r}
library(tidyverse)
library(rcis)
data(scorecard)
glimpse(scorecard)
```

---

## The scorecard dataset: from question to graph

**Question: "Which type of college has the highest average SAT score?"**

When you see a question like this, use this method:

1. identify the variable(s) you need
1. find them in the data
1. check the variable(s) types (categorical or continuous?)
1. use this info to choose the most appropriate plot 
1. *only at this step* write the code for the plot 
1. interpret the graph

Let's try this out! 

<!--
- identify the variable(s) you need: college type and SAT scores

- find them in the data: type glimpse(scorecard) to find out we need type and satavg 

- check the variable(s) types: type is fct (categorical) and satavg is dbl (continuous, wit decimals)

- use this info to choose the most appropriate plot: boxplot or faceted histogram 

- *only at this step* write the code 

- interpret the graph

-->

---

## Which type of college has the highest average SAT score?

#### Option 1: answer using a boxplot

```r
ggplot(data = scorecard, mapping = aes(x = type, y = satavg)) +
  geom_boxplot()
```

```{r}
ggplot(data = scorecard, mapping = aes(x = type, y = satavg)) +
  geom_boxplot()
```

#### [What's a boxplot?](https://datatab.net/tutorial/box-plot){target="_blank"}


---

## Which type of college has the highest average SAT score?

#### Option 2: answer using a faceded histogram

```r
ggplot(data = scorecard, mapping = aes(x = satavg)) +
  geom_histogram() + 
  facet_wrap(facets = vars(type))
```

```{r}
ggplot(data = scorecard, mapping = aes(x = satavg)) +
  geom_histogram() + 
  facet_wrap(facets = vars(type))
```

#### [What's a histogram?](https://datatab.net/tutorial/charts){target="_blank"}

---

## Which type of college has the highest average SAT score?

Now it's time to interpret these graphs:

::: {.callout-note title="Interpreting plots is as important as writing code"}

What do these two graphs we gerated reveal about average SAT scores by type of college? 

Do they tell the same story? Kind of... 

:::

---

## Which type of college has the highest average SAT score?

::: {.callout-note title="Interpreting plots is as important as writing code"}

**Main takeaway:**

Both plots show that private nonprofit schools tend to have the highest SAT scores, followed by public schools. Private for-profit schools tend to have lower SAT scores.

**But...**

The *boxplot* makes it easier to compare the typical SAT score and the spread of SAT scores across school types.

The *faceted histogram* shows the full distribution within each school type. It also makes it easier to see that some groups have many more schools than others. This difference in sample size isn’t visible in the boxplot but is important.

**Why use both?**

The two graphs tell a similar story, but they highlight different things. The boxplot is better for comparing groups quickly. The histogram gives a fuller picture of the distribution and reminds us that group size matters.

:::

<!--
## Asking More Questions with dplyr (sneak peek for next week!)

How many schools are in **each type?** 
```r
scorecard %>% count(type) 
```

Which schools are categorized as **private for-profit?**
```r
scorecard %>%
  filter(type == "Private, for-profit") %>%
  select(name, state, type, satavg, cost)
```

What about the **University of Chicago?**
```r
scorecard %>% filter(name == "University of Chicago")
```

Etc. 

The code below creates a summary dataset with `type` of university as rows and the `satavg` mean as a column:

```{r}
scorecard %>% 
  group_by(type) %>%
  summarize(mean_type = mean(satavg, na.rm = TRUE))
```
-->

---

## Once you draw a plot, ask yourself:

::: columns
::: column

#### Substantive questions:
- What does this graph tell?
- Are there patterns? Outliers?
- What hypotheses can I generate?
- What else I want to know?
- Should I dig deeper with `dplyr`?
- Should I wrangle my data and plot again?
- Etc.

:::
::: column

#### Stylistic questions:
- Is the chosen plot appropriate (matches variable type)?
- Is the plot clear and easy to understand?
- Is it too busy or too simple?
- Could it be improved? (titles, labels, colors, etc.)
- Can I make an additional plot with these variables?
- Etc.

:::
:::

# 2. Match Graph Types to Variable Types {.slide .center .middle}

---

## Types of Visualizations and Best Graph Types

### 1. Which kind of variation do I want to represent?

- **Univariate:** within one single variable
- **Bivariate:** among two variables
- **Multivariate:** among three variables

<br>

### 2. What type(s) are my variables?

- **Continuous** (e.g., `satavg` average SAT score, `cost` published cost)
- **Categorical** (e.g., `type` school type, `state` U.S. state)
- Other types (often treated as categorical): ordinal, nominal, binary

</br>

*Tip: Ask yourself these two questions before jumping into the code!*

---

## Univariate (One Variable)

**To show how values vary within a single variable**:

- One **continuous** variable → **Histogram**
- One **categorical** variable → **Bar Chart**

<br>

::: {.left}

| Function           | What Gets Counted                                 | When to Use             |
|--------------------|---------------------------------------------------|--------------------------|
| `geom_histogram()` | How many values fall into each numeric bin        | When `x` is continuous   |
| `geom_bar()`       | How many observations are in each category        | When `x` is categorical  |

:::

</br>

---

## Bivariate (Two Variables)

**To show how two variables co-vary**:

- Two **continuous** variables → **Scatterplot**
- One **categorical** + one **continuous** → **Box Plot**
- Two **categorical** variables → **Grouped or Stacked Bar Chart**

---

## Multivariate (Three Variables)

**To compare patterns across subgroups**:

- One **categorical** + two **continuous** → **Faceted Scatterplot**
- Two **categorical** + one **continuous** → **Grouped Box Plot**


---

## Main Graph Types and When to Use Them

| Type of Graph        | `ggplot2` Function                                                              | Input Variables                                                                 | Goal                                  |
|----------------------|----------------------------------------------------------------------------------------|----------------------------------------------------------------------------------|----------------------------------------------------------|
| Histogram            | [geom_histogram()](https://ggplot2.tidyverse.org/reference/geom_histogram.html)       | One **continuous** (e.g., age, income)                              | Show the **distribution** of values                     |
| Bar Chart            | [geom_bar()](https://ggplot2.tidyverse.org/reference/geom_bar.html)                   | One **categorical** (e.g., region, gender)                           | Show **frequencies** or **counts** of categories        |
| Stacked Bar Chart    | [geom_bar() + fill](https://ggplot2.tidyverse.org/reference/geom_bar.html)            | Two **categorical** variables (e.g., region by gender)                         | Compare **parts of a whole** across categories          |
| Scatterplot          | [geom_point()](https://ggplot2.tidyverse.org/reference/geom_point.html)               | Two **continuous** variables (e.g., height vs. weight, price vs. rating)        | Show **relationship** or **correlation**                |
| Box Plot             | [geom_boxplot()](https://ggplot2.tidyverse.org/reference/geom_boxplot.html)           | One **continuous** + one **categorical** (e.g., income by gender)           | Compare **distributions**, spot **outliers**            |
| Faceted Scatterplot  | [facet_wrap() + geom_point()](https://ggplot2.tidyverse.org/reference/facet_wrap.html) | Two **continuous** + one **categorical** (e.g., by country or year)             | Compare **patterns across groups**                      |

*These are the most common combinations. There are more options as you explore deeper!*


# 3. Practice {.slide .center .middle}

---

## 💻 Practice: Make Graphs!

On the next slide, you will see list of graphs to create. 

Use the scorecard dataset to choose and create the most appropriate graph for each prompt.

- **BEFORE plotting:** Identify the variables you need. Then ask: are they categorical, continuous, or both? Use these slides as a guide!

- **WHILE plotting:** Keep the graph simple. Focus on the basic code; do not add labels, colors, or other customizations yet. Use the documentation when needed: learning how to look things up is an important skill.

- **AFTER plotting:** Stare at your graph... What patterns, differences, outliers, or surprises do you notice? What does the graph tell us about the data?

---

## 💻 Practice

<!-- For regular CSS class, add some dpyr commands in task and also bring back TASK 4  Display the total number of schools in the U.S. by school type (n = 3) and by state (n = 54). -->

Write the code needed to generate the graphs below using what we have learned so far. The graphs will not look polished yet, and that is okay. The most important goal is to understand the basic structure of the code. Customization can always be added later.

Share your solutions to each graph here <https://codeshare.io/5z6PkW>

**Graph 1:** Plot the total number of schools by school type (`type`)

**Graph 2:** Plot the annual total cost of school attendance (`cost`)

**Graph 3:** Plot the annual total cost and net cost of attendance to schools in the U.S. (`cost`, `netcost`)

**Graph 4:** Plot the annual total cost of attendance by school type (`cost`, `type`)

**Graph 5:** Plot the annual total cost of attendance and net cost of attendance by school type (`cost`, `netcost`, `type`)

---

## 💻 Download Today Materials

Download today’s class materials from our website for: 

- further insights into these tasks
- additional practice exercises (optional)

---

## Takeaways and Tips

Follow this approach to move from quick exploration to polished, communicative plots:

::: columns
::: column

#### When exploring your data:
- Start with simple and plots to get a sense of the distribution and relationships
- Focus on interpretation first, not style
- Ask questions: What stands out? Are there outliers, trends, or surprises? What else do you want to know?

:::
::: column

#### Once you settle on a plot:
- Refine the code to improve readability and aesthetics
- Add styling elements (labels, scales, legends, and themes. etc. as needed)
- Use polished plots for assignments, reports, or presentations

:::
:::

*Typically, researchers create many exploratory plots and only a few make it to the final report.*

---

## Recap: What We Learned Today

- Pick graph types that match the types of variables in your data
- Practice using `ggplot2`
- Read and interpret visualizations

---

## To print these slides as pdf

Click on the icon bottom-left corner (the three lines) \> Tools \> PDF Export Mode \> Print as a Pdf

