---
title: "Computing for the Social Sciences: Lecture 11"
subtitle: "Topics: Conditional Statements with if, else, and else if"
author: "Sabrina Nardin, Fall 2025"
format:
  revealjs:
    theme: simple
    slide-number: true
    incremental: false
    css: ../style/styles.css
    code-overflow: wrap
    highlight-style: github
    ratio: 16:9
    chalkboard: true
    toc: false
    center: true
---


# Agenda {.center}

::: {.agenda-list}
1. Overview of Control Structures: Conditional Statements and Loops
2. Conditional Statements
:::

<span style="font-size: 0.8em; color: #666;">
*Slides last updated on **November 7, 2025**. Slides authored by Sabrina Nardin. AI used to polish slides style and fix typos.*
</span>

<!--
<span style="font-size: 0.8em; color: #666;">
*These slides were last updated on **`r format(Sys.Date(), '%B %d, %Y')`***
</span>
-->

<!--
Tips for Family Program Class

Prep
- Print warm up (25 copies?)
- Bring papers and markers 

Intro
- Start by introducing yourself
- Instructor in MACSS Program, which is about
- This is a lecture from a class called "CSS" 
- Targets mix of undergraduate, PhD and MA students
- They typically enroll because...
- Most students are beginners
- Show website, schedule, etc.

Start lecture, show slides, say
- By the time they arrive to this lecture they have learned...
- We run this class and every class interacively (students bring their laptop and have access to R and code with me or do exercises diretly on R) -- we adjust this for today!
- I am going to make you work a bit to give you a flavor of what classes look like, but also sollicit questions on anything unclear as I tell my students no stupid questions at all

Remember
- For each slide, check what audience might not know
- All weird R signs 
- dplyr syntax at point 4
- Explain workbench (R interface) as soon as you log into it
-->


# 1. Overview of Control Structures {.slide .center .middle}

---

## Control Structures: What's the Idea?

Up to this point, our R code has been running line by line following in a fixed sequence.

<br>

::: {.callout-tip title="**Now we ask: Can our code be dynamic vs fixed?**"}

Example: Can our code skip some lines and run others depending on a condition? Can it repeat itself when needed?

Yes, with control structures!
:::

</br>

---

## What Are Control Structures?

Control structures let us write code that can:

- **Make decisions** → with **conditional statements** like `if`, `if...else`, etc. (today's topic)

- **Repeat actions**  → with **loops** like `for`, `while` (next lecture topic)

Conditional statements and loops (e.g., control structures) let our code responds dynamically to different situations!

<!-- 
Make decisions: the code uses conditions, evaluates them, and chooses what to do  
Repeat actions: the code runs certain lines multiple times when needed  

*Today: Conditional Statements and Next time: Loops*
-->

# 2. Conditional Statements  {.slide .center .middle}

---

## Conditional Statements

**Conditional statements translate into programming language our “if this, then that” logic**: the code checks conditions, evaluates them, and acts accordingly.

1. Single test with `if... else`

2. Multiple tests with `if... else if... else`

3. Nested tests by placing one `if` inside another

4. Vectorized tests with `ifelse()` to apply a condition to each element of a vector (not technically a conditional statement like the previous three, but more on this when we get there!)


<!--
## Loops

**Loops let our code repeat actions by running certain lines multiple times when needed**

1. For Loop
2. While Loops
-->

---

## 💻 Practice

1. Download today's class materials from the course website

2. Open the file called `if-else-warm-up.R`

3. Before running the code: look over the syntax and try to predict what the code will output

4. Run the code and compare it to your prediction

5. Share your questions about this code on [Ed Discussion](https://edstem.org/us/courses/86758/discussion)

---

## 1. Single test with "if... else"

Use `if...else` when your code needs to test **one condition** and take one of two possible actions, depending on whether the condition is `TRUE` or `FALSE`.

Syntax:
```r
if (condition) {
    action1 # action performed when condition is TRUE
} else {
    action2 # action performed when condition is FALSE
}
```

Example:
```{r echo = TRUE, results = "hide"}
age <- 14

if (age > 16) {
  print("You can get a driving license")
} else {
  print("You cannot drive")
}
```

---

## 1. Single test with "if... else"

::: {.callout-tip title="Style tip"}

Do you need the space and `{}`?

Technically, the code works without them, like this: 

```r
age <- 14

if (age > 16) print("You can get a driving license") else print("You cannot drive")
```

But omitting spaces and `{}` is considered **bad style** as it hurts readability and makes errors more likely, especially for more complex statements!

:::

---

## 1. Single test with "if... else"

Another example: What's the output of this code?

```{r, echo = TRUE, results = "hide"}
movie_rating <- c(4.5, 4.2, 5.0, 3.8, 2.9, 1.7)

movie_rating_avg <- mean(movie_rating)

if (movie_rating_avg >= 4) {
  print("This movie is great!")
} else {
  print("Not a favorite.")
}
```

Note conditional statements evaluate only one single element (in this example the mean of the vector movie_rating)

<!--
wont work to evaluate each element of the vector, you ll have to put it into a loop, or use a logical vector first and replace
-->

---

## 2. Multiple tests with "if...", "else if...", "else"

Use `if...else if...else` when your code needs to test **multiple conditions** and take different actions depending on which condition is `TRUE`. To do so, ddd one or more **`else if`** statement(s) between the initial `if` and the final `else`

Syntax:
```r
if (condition1) {
    action1   # action performed when condition 1 is TRUE
} else if (condition2) {
    action2   # action performed when condition 2 is TRUE
} else {
    action3   # action performed when conditions 1 and 2 are FALSE
}
```

Example:

```{r, echo = TRUE, results = "hide"}
x <- 0

if (x > 0) {
  print("x is positive")
} else if (x < 0) {
  print("x is negative")
} else {
  print("x is zero")
}
```

---

## 2. Multiple tests with "if...", "else if...", "else"

Another example with multiple `else if` statements:

```r
# Enter a temperature when prompted
temperature <- as.integer(readline(prompt = "Enter today's temperature in Celsius: "))

# Determine the variable weather based on temperature
if (temperature >= 30) {
  weather <- "Hot"
} else if (temperature >= 20) {
  weather <- "Cool"
} else if (temperature >= 10) {
  weather <- "Breezy"
} else {
  weather <- "Freezing"
}
print(weather)
```

---

## Note: with conditional statements sequence matters!

::: {.callout-warning title="Order matters"}
Conditional statements are evaluated in order, so the sequence of your code matters:  

- R checks each condition one by one, from top to bottom  
- R stops checking once it finds a condition that is `TRUE`: it runs that block and skips the rest
- As a rule of thumb: order your conditions from most specific to most general, ending with a default case `else`
:::


---

### 💻 Practice 1

What happens if the conditions are out of order? 

Copy the code below in your console and move the line `temperature >= 10` before the line `temperature >= 20`. Leave the input temperature set as 25. What do you notice?

```r
temperature <- 25

if (temperature >= 30) {
  weather <- "Hot"
} else if (temperature >= 20) {
  weather <- "Cool"
} else if (temperature >= 10) {
  weather <- "Breezy"
} else {
  weather <- "Freezing"
}
print(weather)
```

Once done, [share your code here](https://codeshare.io/G8nxLD){target="_blank"}. Questions? post on [Ed Discussion](https://edstem.org/us/courses/86758/discussion) or ask us directly!
 
<!--
```r
temperature <- as.integer(readline(prompt = "Enter today's temperature in Celsius: "))

if (temperature >= 30) {
  weather <- "Hot"
} else if (temperature >= 10) {   # moved up
  weather <- "Breezy"
} else if (temperature >= 20) {
  weather <- "Cool"
} else {
  weather <- "Freezing"
}
print(weather)
```

If temperature is 25, R hits temperature >= 10 first (which is TRUE) and never reaches temperature >= 20. So it assigns "Breezy" instead than "Cool".
-->

---

### 💻 Practice 2

Copy the code below into your console and add another `else if` statement that checks whether the temperature is between 0 (inclusive) and 10 (exclusive). If that condition is true, set weather to "Cold" (so the “Cold” range is 0–9°C). Leave the input temperature set as 4.

```r
temperature <- 4

if (temperature >= 30) {
  weather <- "Hot"
} else if (temperature >= 20) {
  weather <- "Cool"
} else if (temperature >= 10) {
  weather <- "Breezy"
} else {
  weather <- "Freezing"
}
print(weather)
```

Once done, [share your code here](https://codeshare.io/G8nxLD){target="_blank"}. Questions? post on [Ed Discussion](https://edstem.org/us/courses/86758/discussion) or ask us directly!
 
<!--
```r
# Take user input
temperature <- as.integer(readline(prompt = "Enter today's temperature in Celsius: "))

# Determine weather based on temperature
if (temperature >= 30) {
  weather <- "Hot"
} else if (temperature >= 20) {
  weather <- "Cool"
} else if (temperature >= 10) {
  weather <- "Breezy"
  
  # NEW CODE
} else if (temperature >= 0 & temperature < 10) {  
  weather <- "Cold"
  
} else {
  weather <- "Freezing"
}
print(weather)
```

Takeaway: 

Use AND (&) when both conditions must be TRUE (here, >= 0 and < 10) at the same time for the code inside the {} runs

Here temperature must be greater than or equal to 0, and temperature must be less than 10. That’s what defines our “Cold” range (0–9°C).

If you write OR only one must be TRUE, for the code to run. 
So this condition is TOO BROAD would always be true, because every number is either ≥ 0 or < 10.

HERE YOU WON'T NOTICE IT THOUGH, BUT  USING OR IS STILL LOGICALLY INCORRECT

-->

---

### 💻 Practice 3

When do we use `&` (and) versus `|` (or) in conditional statements? Try these two examples, what do you notice and why?

```{r, echo = TRUE, results = "hide"}
# Set temperature (25°C, which equals 77°F)
temperature <- 25

# Example 1
if (temperature >= 0 & temperature < 10) {
  weather <- "Cold"
} else {
  weather <- "Warm"
}
print(weather)

# Example 2
if (temperature >= 0 | temperature < 10) {
  weather <- "Cold"
} else {
  weather <- "Warm"
}
print(weather)
```

<!--
Use AND (&) when both conditions must be TRUE (here, >= 0 and < 10) at the same time for the code inside the {} runs

Here temperature must be greater than or equal to 0, and temperature must be less than 10. That’s what defines our “Cold” range (0–9°C).

If you write OR only one must be TRUE, for the code to run. 
So this condition is TOO BROAD would always be true, because every number is either ≥ 0 or < 10.
-->

---

## 3. Nested tests by placing one "if" inside another

<!-- The %% modulo operator calculates the remainder from a division operation. Here 15 : 2 = 7, 1 is the reminder, thus the number is odd -->

Use nested `if...else` statements when your code needs to check **conditions within conditions**. This allows you to make a second decision **only after** a first condition is met.

Example (the modulo `%%` returns the remainder after a division):

```{r, echo = TRUE, results = "hide"}
x <- 15

if (x > 0) {
  # Case 1: x is positive
  if (x %% 2 == 0) {
    print("x is a positive even number")
  } else {
    print("x is a positive odd number")
  }

} else {
  # Case 2: x is negative
  if (x %% 2 == 0) {
    print("x is a negative even number")
  } else {
    print("x is a negative odd number")
  }
}
```

What output do you expect if `x <- -6` and if `x <- 0`? 

<!-- current places 0 as negative even number but incorrect, so we need to improve the code! -->

---

## 3. Nested tests by placing one "if" inside another

Same example, but this checks also the condition `x <- 0`

```{r, echo = TRUE, results = "hide"}
x <- 0

if (x > 0) {
  # Case 1: x is positive
  if (x %% 2 == 0) {
    print("x is a positive even number")
  } else {
    print("x is a positive odd number")
  }
} else if (x < 0) {
  # Case 2: x is negative
  if (x %% 2 == 0) {
    print("x is a negative even number")
  } else {
    print("x is a negative odd number")
  }
} else {
  # Case 3: x is exactly zero
  print("x is zero")
}
```

---

## 4. Vectorized tests with ifelse() 

Our final case for today is `ifelse()`: it looks like a conditional statement, but it's actually a function that applies a condition to **each element of a vector**. 

Syntax:
```r
ifelse (condition to be evaluated,
        action performed when condition is TRUE,
        action performed when condition is FALSE)
```

Example:
```{r, echo = TRUE, results = "hide"}
ifelse(sqrt(16) > 3, 
       sqrt(16),
       0)
```

<!--
if_else(): more flexibile and handles complex conditional logic; returns a scalar

ifelse(): simpler and more suited for simple assignments or to transform variables in a dataframe; returns a vector
-->

---

## 4. Vectorized tests with ifelse() 

Another example, this time with a vector as input. What is the output of this code?

```{r, echo = TRUE, results = "hide"}
numbers <- c(10, 6, 7)
ifelse(numbers %% 2 == 1, 
       "odd",
       "even")
```

<br>

::: {.callout-note title="How ifelse() works"}

- If the input of ifelse() is a vector, the output is also a vector: **ifelse() evaluates each element of the vector** and applies the corresponding action

- This works because **ifelse() supports vectorized operations**: operations directly applied on entire vectors, rather than looping through individual elements one-by-one

:::

<!--
How can we rewrite the same code using the standard `if-else` statement?

```
# won't work
numbers <- c(10, 6, 7)
if (numbers %% 2 == 1) { 
  print("odd")
} else {
  print("even")
}
```

```
# need a loop
numbers <- c(10, 6, 7)
for (i in 1:length(numbers)) {
  if (numbers[i] %% 2 == 1) {
    print("odd")
    } 
else 
  print("even")
 }
```
-->

---

## 4. Vectorized tests with ifelse() 

:::: {.columns}
::: {.column width="50%"}
### Regular if... else statements
- Can only test **one value at a time** 
- Think of them as asking **one yes/no question to one person at a time**
:::

::: {.column width="50%"}
### ifelse()
- Can test a **whole vector at once**
- Think of it as asking **the same question to everyone in a group at once**
:::
:::: 

Examples:
```r
x <- c(5, 10, 15)

# Using if / else: works only with a single value
if (x > 10) {
  "High"
} else {
  "Low"
}

# Using ifelse(): works element-wise (vectorized)
ifelse(x > 10, "High", "Low")
```

---

## 4. Vectorized tests with ifelse() 

For this reason, `ifelse()` is often used to make changes in a data frame. Example: what is the output of this code?

```{r, echo = TRUE, results = "hide"}
library(tidyverse)

qualify <- tibble("Athlet" = c("Noah", "Julio", "Nick", "Maria"), 
                  "Scores" = c(32, 37, 28, 30))

# using base R 
ifelse(qualify$Scores > 30, 
       "Admitted", 
       "Rejected")

# same using dplyr
qualify %>%
  mutate(status = ifelse(Scores > 30,
                       "Admitted",
                       "Rejected"))
```

<!--  FOR CSS CLASS IF THIS COMES AFTER DATA STRUCTURES, ADD THIS SLIDE IN
AND EXPLICITLY CONNECT IT BACK TO PREVOIUS CLASS 

You can also use double square bracket to access column Scores by position vs name 

# using base R to access column Scores by position
# double square brackets on a df return a vector
ifelse(qualify[[2]] > 30, 
       "Admitted", 
       "Rejected")

# single square brackets on a df return a tibble
ifelse(qualify[2] > 30, 
       "Admitted", 
       "Rejected")


# stick with [[]] for ector operations, safer always returns a vector
-->

---

## 4. Vectorized tests with ifelse() 

In the previous example, we only had two possible categories: "Admitted" and "Rejected"

What happens to our code if we add a third category, say "Waitlisted"? We can write a **nested `ifelse()`** like this:

```{r, echo = TRUE, results = "hide"}

qualify <- tibble("Athlet" = c("Noah", "Julio", "Nick", "Maria"), 
                  "Scores" = c(32, 37, 28, 30))

qualify <- tibble(
  Athlet = c("Noah", "Julio", "Nick", "Maria"), 
  Scores = c(32, 37, 28, 30))

qualify %>%
  mutate(Status = ifelse(Scores > 35,            # First condition
                         "Admitted",             # If TRUE, assign "Admitted"
                         ifelse(Scores >= 30,    # Second condition (checked only if first is FALSE)
                                "Waitlisted",    # If TRUE, assign "Waitlisted"
                                "Rejected")))    # If both conditions are FALSE, assign "Rejected"

```

<!--
# more formally...

ifelse(condition1,
       "Label1",           # used if condition1 is TRUE
       ifelse(condition2,
              "Label2",    # used if condition1 is FALSE and condition2 is TRUE
              "Label3"))   # used if BOTH condition1 and condition2 are FALSE
-->

---

## 4. Vectorized tests with ifelse() 

Nested `ifelse()` statements work, but they make the code harder to read. 

Compare the previous example with this version that avoids nesting:

```{r, echo = TRUE, results = "hide"}

qualify <- tibble("Athlet" = c("Noah", "Julio", "Nick", "Maria"), 
                  "Scores" = c(32, 37, 28, 30))

qualify %>%
  mutate(Status = "Rejected") %>%                                  # Start with a default label
  mutate(Status = ifelse(Scores >= 30, "Waitlisted", Status)) %>%  # Update if the condition is met
  mutate(Status = ifelse(Scores > 35, "Admitted", Status))         # Update again if the higher condition met
```

<!--
# or can also start from Admitted and down

qualify %>%
  mutate(Status = "Admitted") %>%                                   # start with highest category
  mutate(Status = ifelse(Scores <= 30, "Waitlisted", Status)) %>%  # downgrade if below admission threshold
  mutate(Status = ifelse(Scores < 25, "Rejected", Status))          # downgrade again if below waitlist threshold
-->

<br>

::: {.callout-tip title="Tip"}
When you have 2+ conditions: use separate `mutate()` + `ifelse()` steps (as shown here), instead of nesting.
:::


</br>

<!--
## When to Use `ifelse()` Instead of `case_when()`

It depends but as a rule of thumb...

Use `ifelse()` when:

- You're checking **only one condition**
- You want a **simple inline transformation**  
  (e.g., `ifelse(x > 0, x * 2, x)`)
- You're assigning `NA` for some values  
  (e.g., `ifelse(x < 0, NA, x)`)

Use `case_when()` when:

- You have **3 or more conditions**
- You want better readability for **categorical recoding**

-->

---

## 4. Vectorized tests with ifelse() 

People often use `ifelse()` to recode variables in a dataframe. 

For example, the variable `decisionDirection` takes four values: 1 = Conservative, 2 = Liberal, 3 = Other, and NA.  
Recode it to three values: 0 = Conservative, 1 = Liberal, and NA for both NA and Other.  

```r
data |>
  mutate(decisionDirection = ifelse(
    decisionDirection == 3, # condition to evaluate
    NA,                     # if true, set to NA
    decisionDirection - 1   # if false, subtract one to value
    ))
```

<br>

::: {.callout-tip title="Tip"}
Tip: Prefer `case_when()` over `ifelse()` for recoding variables (see the recoding class): it is easier to read and scale. 
:::

</br>



<!--
More on this:
https://docs.ycrc.yale.edu/r-novice-gapminder/07-control-flow/
Comparing ifelse() and for loop:
https://www.r-bloggers.com/2020/02/if-else-and-ifelse/
-->

---

## 4. Vectorized tests with ifelse() 

**Note there is also a tidyverse version of the base R `ifelse()` function!**

The tidyverse version is called `if_else()` <https://dplyr.tidyverse.org/reference/if_else.html>

The two are very similar, hence I introduced only on here, pick the one you prefer.

---

## 💻 Practice

More practice exercises in today's class materials (downloaded from the website)

---

## Recap: What We Learned Today

- Control structures let us change the flow of execution in our code  
- We focused on conditional statements: how to make decisions using `if`, `else`, and `else if`
- We practiced different types of conditionals: single test, multiple tests, and nested tests  
- Run vectorized tests with the function `ifelse()` and how this differs from standard conditional statements


---

## To print these slides as pdf

Click on the icon bottom-right corner \> Tools \> PDF Export Mode \> Print as a Pdf

