---
title: "Computing for the Social Sciences: Lecture 12"
subtitle: "Topics: Strings and Regex with stringr; Dates with lubridate"
author: "Sabrina Nardin, Summer 2026"
format:
  revealjs:
    theme: simple
    slide-number: true
    incremental: false
    css: ../style/styles.css
    code-overflow: wrap
    highlight-style: github
    ratio: 16:9
    chalkboard: true
    toc: false
    center: true
---


# Agenda {.center}

::: {.agenda-list}
1. Strings and Regular Expressions with stringr
2. Manipulating Dates with lubridate  
:::

<!--
### Slides are in progress and will be updated before class.

::: {.content-hidden}
-->

# 1. Strings and Regular Expressions with stringr {.inverse .middle}

--- 

## Introduction

These are strings: 

```r
string1 <- "This is a string"

string2 <- 'If I want to include a "quote" inside a string, I use single quotes'
```

Today we focus on methods for manipulating strings: regular expressions!

<!-- Not 1:1 allignment but haver focus on Reg Expr because super important to know them! 
So formally a string is... -->

--- 

## String and Character Vector

A **string** is a sequence of characters (a piece of text) enclosed in either double `"` or single quotes `'`. Strings can include letters, numbers, symbols, and whitespace characters.

```r
string <- "Ciao, my name is Sabrina and I am 99 years old!!"
class(string)
length(string)
```

--- 

## String and Character Vector

A **character vector** is a collection of strings: each element of the vector is a string! 

```r
char_vect <- c("Ciao come stai?", "Ciao", "Hello", "etc.") 
class(char_vect)
length(char_vect)
```

Thus single string, like in the previous examples, is also a character vector of length one.

We use character vectors when we need to handle multiple pieces of text together (e.g., when we handle multiple strings vs. only one).

---

## Regular Expressions: definitions and uses

Regular expressions ("regex" or "regexes") **are strings containing normal characters and/or special meta-characters.** They describe a specific pattern to match in a given text.

More formally, regex:
* is a *language on its own right* 
* used for *pattern matching* 
* adopted by many programming languages, such as R, Python, and others!

Example:

```r
email_pattern <- "[a-zA-Z0-9._%+-]+@[a-zA-Z0-9.-]+\\.[a-zA-Z]{2,}"
```

<!--
But then you might ask yourself...
given our ability to manipulate strings and our ability to test for equivalence (with the == operator) 
or to test whether some string contains another (with the %in% operator), we don't technically need regular expressions... Not exactly! We want to use regex becasue they are POWERFUL
-->

---

## Regular Expressions: definitions and uses

Regular expressions are powerful! They allow us to find patterns **in any task that deals with text,** such as NLP (Natural Language Processing) or data-cleaning tasks that involve text. 

For example, you can use regular expressions to:

* Extract specific elements from texts (e.g., find words that include a given set of letters, find all past tenses in a text, etc.)
* Perform textual substitutions (e.g. find and replace HTML tags after you scraped a page)
* Verify text format (e.g. "Is this email/phone-number/address valid?")


---

## The stringr() package in R

Today we learn about the package `stringr` and regular expressions!

When you use regular expressions, most likely you will need to use your them together with one of the functions from the `stringr()` package. 

This package includes several functions that let you: detect matches in a string, count the number of matches, extract them, replace them with other values, split a string based on a match, etc. 

---

## The stringr() package in R

Fundamental `stringr()` functions and their use:

* `str_view()` return the first regex match
* `str_view_all()` return all regex matches (deprecated in the last version of `stringr`)
* `str_detect()`: detect matches in a string
* `str_count()`: count the number of matches
* `str_extract()` and `str_extract_all()`: extract matches
* `str_replace()` and `str_replace_all()`: replace matches
* `str_split()`: split a string based on a match

---

## Resources

* [`stringr()` documentation and cheatsheet](https://stringr.tidyverse.org/)

* [Chapter 15](https://plsc-31101.github.io/course/strings-and-regular-expressions.html#applying-regex) by Rochelle Terman, explains `stringr()` (using the *R for Data Science* textbook)  

* Read [Chapter 14 "Strings"](https://r4ds.hadley.nz/strings) in "R for Data Science" 2nd Edition (read it all, but especially 14.4 "Extracting data from strings").

* Read [Chapter 17](https://bookdown.org/rdpeng/rprogdatascience/regular-expressions.html#the-stringr-package) from *R Programming for Data Science*. This book covers the entire range of regular expressions packages and functions: you do not need to understand everything, focus on the big picture. In-class we talk about `stringr()`

* For more in-depth info on regular expressions check [this excellent tutorial](https://github.com/ziishaned/learn-regex/blob/master/README.md)

---

## 💻 Practice {.inverse .middle}

Download today's class material from the website to learn more about regular expression and `stringr()`

# 2. Manipulating Dates with Lubridate {.slide .center .middle}

---

## Date and Times with Lubridate

Lubridate (also part of the `tidyverse`) is a handy package that makes it easier work with dates & times <https://r4ds.hadley.nz/datetimes.html>

::: {.fragment}

### Common Lubridate functions:

- Turn strings into dates or date-times with `ymd()`, `mdy()`, `dmy()`, `ymd_hms()`  
- Turn variables into dates or date-times with `make_datetime()`  
- Extract individual date components with `year()`, `month()`, `day()`, `wday()`  
- Round the date to a nearby unit of time `+ days(1)`, `floor_date()`, `round_date*()`  
- Handle time spans (multiple options, see Ch. 17)
- Handle time zones `with_tz()`, `force_tz()`  
:::

---

## Turn strings into dates or date-times

```{r, echo = TRUE}
library(lubridate)
ymd("2017-01-31")
mdy("January 31st, 2017")
dmy("31-Jan-2017")
ymd_hms("2017-01-31 20:11:59")
```

---

## Turn variables into dates or date-times

::: {.columns}
::: {.column width="40%"}
```{r, echo = TRUE}
library(tidyverse)
library(nycflights13)
flights |> 
  select(year, month, day, hour, minute)
```
:::

::: {.column width="60%"}
```{r, echo = TRUE}
# create a date/time from this data with make_datetime()
flights |> 
  select(year, month, day, hour, minute) |> 
  mutate(departure = make_datetime(year, month, day, hour, minute))

```
:::
:::

<br>

More examples: <https://r4ds.hadley.nz/datetimes.html#from-individual-components>

</br>

---

## Extract individual date components 

```{r, echo = TRUE}
datetime <- ymd_hms("2025-10-22 12:30:33")

year(datetime)
month(datetime)
mday(datetime)
yday(datetime)
wday(datetime)
```

More examples: <https://r4ds.hadley.nz/datetimes.html#getting-components>

---

##  Round the date to a nearby unit of time 

```{r, echo = TRUE}
# create date object from string
dt <- ymd_hms("2025-10-22 12:45:00")
dt

# round down with floor_date() 
floor_date(dt, "day")
floor_date(dt, "month")

# round up with ceiling_date()
ceiling_date(dt, "day")
ceiling_date(dt, "month")
```

More examples: <https://r4ds.hadley.nz/datetimes.html#rounding>

---

## Handle Time Spans

```{r, echo = TRUE}
# How old are you?
age <- today() - ymd("2001-10-22")
age

as.duration(age)
```

More examples: <https://r4ds.hadley.nz/datetimes.html#time-spans>

---

## Handle Time Zones

```{r, echo = TRUE}
greenwitch <- ymd_hms("2025-10-22 6:45:00")

# display same instant in another time zone
# shows local Central Time equivalent
with_tz(greenwitch, tzone = "America/Chicago")

# assign a time zone (without shifting actual clock time)
# interprets existing time as if it were recorded in Chicago
force_tz(greenwitch, "America/Chicago")
```

More examples: <https://r4ds.hadley.nz/datetimes.html#time-zones>
