---
title: "Computing for the Social Sciences: Lecture 13"
subtitle: "Topics: Control Flow 2 (cont.); Functions 1"
author: "Sabrina Nardin, Fall 2025"
format:
  revealjs:
    theme: simple
    slide-number: true
    incremental: false
    css: ../style/styles.css
    code-overflow: wrap
    highlight-style: github
    ratio: 16:9
    chalkboard: true
    toc: false
    center: true
---


# Agenda {.center}

::: {.agenda-list}
1. Using For Loops with Dataframes 
2. Alternatives to For Loops in R
3. While Loops
4. Introduction to Functions
:::

<span style="font-size: 0.8em; color: #666;">
*Slides last updated on **November 11, 2025**. Slides authored by Sabrina Nardin. AI used to polish slides style and fix typos.*
</span>

# 1. Using For Loops with Dataframes {.slide .center .middle}

---

## Same Task With and Without a For Loop

```{r echo = TRUE, results = "hide"}

library(tidyverse)
library(palmerpenguins)
data(penguins)
```

```{r}
glimpse(penguins)
```

<br>

#### Task: Calculate the mean value of several columns in this data

</br>

---

## Calculate the mean value of several columns *without a loop*

We can take the `mean()` function, and apply it to each column:

```{r echo = TRUE, results = "hide"}

penguins %>% summarize(avg_bill_length = mean(bill_length_mm, na.rm = TRUE))
penguins %>% summarize(avg_bill_depth = mean(bill_depth_mm, na.rm = TRUE))
penguins %>% summarize(avg_flipper_length = mean(flipper_length_mm, na.rm = TRUE))
penguins %>% summarize(avg_body_mass = mean(body_mass_g, na.rm = TRUE))
                       
```

<!--
```r
mean(penguins_clean$bill_length_mm)
mean(penguins_clean$bill_depth_mm)
mean(penguins_clean$flipper_length_mm)
mean(penguins_clean$body_mass_g)
```
-->

This works, but requires a lot of copy/paste! How can we do the same thing with a for loop?

---

## Automate the same task *with a loop*

First, initialize an empty vector to store the results. 

Second, use a for loop to calculate the mean of each (numeric) column of this penguins dataframe. 

Let's run this code:

```{r echo = TRUE, results = "hide"}

output <- vector(mode = "double", length = ncol(penguins))

for (i in seq_along(penguins)) {
  output[i] <- mean(penguins[[i]], na.rm = TRUE)
}

output
```

---

## Why output[i] and penguins[[i]]?

```{r echo = TRUE, results = "hide"}

output <- vector(mode = "double", length = ncol(penguins))

for (i in seq_along(penguins)) {
  output[i] <- mean(penguins[[i]], na.rm = TRUE)
}

output
```

::: {.callout}

**`output` is a vector where we store the loop results:**

- To access elements of a vector, use single square brackets `[]` 
- So `output[i]` assigns the computed value into the *i-th position* of that vector 

**`penguins` is a dataframe (or tibble):**

- But `mean()` needs a numeric vector  
- To access a dataframe column *as a vector*, use double brackets `[[]]`  
- So `penguins[[i]]` gets the *i-th column* of the dataframe, and turns it into a vector 

:::


---

## Why output[i] and penguins[[i]]?

```{r echo=TRUE, results="hide"}
is.vector(penguins[[3]])           # TRUE (the raw vector)
is.data.frame(penguins[3])         # TRUE (a 1-col tibble/data frame)

mean(penguins[[3]], na.rm = TRUE)  # works, correct type (vector)
mean(penguins[3], na.rm = TRUE)    # error, wrong type (1-col tibble)
```

<br>

::: callout 

**With data frames:** `[[ ]]` returns the column itself as a vector, while `[ ]` returns a one-column data frame or tibble

**With vectors:** `[]` and `[[ ]]` behave the same, but is good practice to consistently use `[]` when subsetting vectors 

*Review Lecture 10 and the Data Structures quiz for details.*

:::

</br>

---

## 💻 Practice Subsetting Data frames 

### Answer these questions:

1. What does `is.vector(penguins[["bill_len"]])` return, and why are there quotation marks around the column name? And what does `penguins[["bill_len"]]` return? 

2. Write code to access the fourth column of the penguins dataset in four ways:

    - By column name, returned as a vector
    - By index, returned as a vector
    - By column name, returned as a dataframe
    - By index, returned as a dataframe

For each object, check its data structure using  `is.vector()` or `is.data.frame()` (which return true or false) or `str()` which provides an overview of the object. 

3. What does `is.vector(penguins[[1]])` return and why the output is not what we expect? What about `is.data.frame(penguins[1])`?

# 2. Alternatives to For Loops in R {.slide .center .middle}

---

## Why We Even Learn Loops in R?

R is a **vectorized language**, meaning it's able to apply operations to entire vectors in one go.

This also implies that many tasks in R can and should be done **without writing a for loop**.

Still, understanding loops is important:

- to build **logic and control flow** skills
- for **non-vectorized tasks**, like row-by-row operations
- to **debug** more complex workflows

<!--
> Even if you don’t use loops every day, knowing how they work will strengthen your ability to write efficient and readable R code.
-->

---

## Three Main Alternatives to For Loops in R

When working with dataframes, R provides the following alternatives to for loop:

| Function Family        | Package     | Best Used For                                  | Docs |
|------------------------|-------------|------------------------------------------------|-------------|
| `map_*()`              | purrr     | Applying functions to elements of lists or columns | [map docs](https://purrr.tidyverse.org/reference/map.html) |
| `across()`             | dplyr     | Applying the same function to multiple columns inside `mutate()` or `summarize()` | [across docs](https://dplyr.tidyverse.org/reference/across.html) |
| `*apply()`             | base R      | Row/column-wise operations on matrices or data frames | [apply docs](https://stat.ethz.ch/R-manual/R-devel/library/base/html/apply.html) |

<br>

The asterisk `*` is a placeholder for a family of related functions (e.g., `map_dbl()`, `map_chr()`, `sapply()`, etc.)

</br>

<!--
1. `map_*()` functions from `purr` <https://purrr.tidyverse.org/reference/map.html>
2. `across()` function from `dplyr` <https://dplyr.tidyverse.org/reference/across.html>
3. `*apply()` functions from base R <https://stat.ethz.ch/R-manual/R-devel/library/base/html/apply.html>

The asterisk `*` is a placeholder indicating a *family* of related functions (e.g., `map_dbl()`, `map_chr()`, `sapply()`, etc.).

- Use the `map_*()` functions from the *purrr* package for flexible iteration  
- Use `across()` with `summarize()` or `mutate()` for column-wise operations  
- Use `apply()` to iterate over rows or columns in a matrix or data frame
-->

---

## Option 1: Replacing for loops with map() functions

<!-- As we have seen in the slides and practice, we have to write a lot of code for a for loop, for example to calculate a straightforward operation like the mean or median etc. so R gives us a shortcut: map functions which come from the purrr package in R and are much more compact (go to the documentation page)
-->

The so-called "map functions" come from the `purr` package in R: <https://purrr.tidyverse.org/reference/map.html>

**There are several `map_*()` functions**  each creates a different type of output (this is the same idea as in the `for loop` when we specify the `mode` of our output vector):

- `map()` makes a list
- `map_lgl()` makes a logical vector
- `map_int()` makes an integer vector
- `map_dbl()` makes a double vector
- `map_chr()` makes a character vector

Let's see a few examples using the `penguins` data.

---

## Option 1: Replacing for loops with map() functions

Pick the appropriate `map()` function (there are several!) and specify at least two main arguments:

- what you are iterating over
- what you are calculating

```{r echo = TRUE, results = "hide"}

penguins %>% 
  select(where(is.numeric)) %>%
  map_dbl(mean, na.rm = TRUE)

```

<!-- ADD PRACTICE EXERCISE
## 💻 Practice
-->

---

## Option 2: Replacing for loops with across()

**Another option to replace a for loop is the `across()` function from `dplyr`**

**What it does:** applies the same operation (e.g., mean) to multiple columns at once  

**Best use:** since it comes from `dplyr`, it works seamlessly inside `dplyr` verbs like `mutate()` and `summarize()` which these are its favorite verbs to work with!

---

## Option 2: Replacing for loops with across()

We looked at this example earlier (which calculates the mean of several columns in a data frame) and we rewrote it using a for loop. 

```{r echo = TRUE, results = "hide"}

penguins %>% summarize(avg_bill_length = mean(bill_length_mm, na.rm = TRUE))
penguins %>% summarize(avg_bill_depth = mean(bill_depth_mm, na.rm = TRUE))
penguins %>% summarize(avg_flipper_length = mean(flipper_length_mm, na.rm = TRUE))
penguins %>% summarize(avg_body_mass = mean(body_mass_g, na.rm = TRUE))
                       
```

Now we rewrite it again but this time using `across()`

---

## Option 2: Replacing for loops with across()

Rewrite previous example using `across()`:

```{r echo = TRUE, results = "hide"}

penguins %>% 
  summarize(
    across(
      .cols = where(is.numeric),           # select columns by type
      .fns = ~ mean(.x, na.rm = TRUE)))    # apply mean, skipping NA
```

<br>

**The function `across()` has two main arguments:**

- **`.cols`**: the columns to operate on. You can select them by position, name, or type - in this example, by type using `where(is.numeric)`

- **`.fns`**: the function, or list of functions, to apply to each column (in this example, we use `mean()` and `.x` is a placeholder for the current column being processed)

You can omit `.cols` and `.fns` if you pass the arguments in the correct order.

</br>

---

## Option 2: Replacing for loops with across()

Add names to the newly computed means:

```{r echo = TRUE, results = "hide"}

penguins %>% 
  summarize(
    across(
      .cols = where(is.numeric),                   
      .fns = ~ mean(.x, na.rm = TRUE),             
      .names = "avg_{.col}"                   
    )
  )
```

<br>

**What does `\"avg_{.col}\"` mean?**

- `"avg_"` is a literal prefix that will be added to each new column name.  
- `{.col}` is a placeholder that will be replaced with the original column name.

So if you're applying `mean()` to a column named `bill_length_mm`, the result will be named `avg_bill_length_mm`

</br>

<!-- FOR CSS CLASS: ADD THOSE EXAMPLES OR MORE + A PRACTICE EXERCISE

## More examples using the [`worldbank` data](https://data.worldbank.org/)

```{r worldbank}
data("worldbank", package = "rcis")
worldbank
```

---

### `summarize()`, `across()`, and `where()`

```{r summarize-across-where}
# use across() with where() to pick variables based on type (e.g. is.numeric(), etc.)
worldbank %>% 
  group_by(country) %>%
  summarize(across(.cols = where(is.numeric), .fns = mean, na.rm = TRUE))
```

```{r summarize-across-compound}

# or pick variables based on type and whose name begins with "perc"
worldbank %>%
  group_by(country) %>%
  summarize(across(
    .cols = where(is.numeric) & starts_with("perc"),
    .fn = mean, na.rm = TRUE
  ))
```


---

### `across()` and `filter()`

To use `across()` with `filter()`, we need an extra step: `if_any()` or `if_all()`

```{r filter-if-any, echo = TRUE, results = "hide"}

# if_any() keeps rows where the predicate is true for at least one column
worldbank %>%
  filter(if_any(everything(), ~ !is.na(.x)))
```

```{r filter-if-all, echo = TRUE, results = "hide"}

# if_all() keeps rows where the predicate is true for all selected columns
worldbank %>%
  filter(if_all(everything(), ~ !is.na(.x)))
```

-->

---

## Option 3: Replacing for loops with apply() functions

**Finally, a third option to replace a for loop in R is using one of the `apply()` functions from base R**

**What it does:** applies the same operation (e.g., mean) to multiple columns at once  

**Best use:** when working outside tidyverse verbs (this is from base R!) and can be applied across rows or columns (most common)

---

## Option 3: Replacing for loops with apply() functions

We looked at this example earlier (which calculates the mean of several columns in a data frame) and rewrote it using a `for` loop. Now we rewrite it again but this time using `apply()`:

```{r echo = TRUE, results = "hide"}

penguins %>% summarize(avg_bill_length = mean(bill_length_mm, na.rm = TRUE))
penguins %>% summarize(avg_bill_depth = mean(bill_depth_mm, na.rm = TRUE))
penguins %>% summarize(avg_flipper_length = mean(flipper_length_mm, na.rm = TRUE))
penguins %>% summarize(avg_body_mass = mean(body_mass_g, na.rm = TRUE))
                       
```

---

## Option 3: Replacing for loops with apply() functions

Rewrite previous example using `apply()`:

```{r echo = TRUE, results = "hide"}

# manually select only numeric column using base R syntax
penguins_numeric <- penguins[ , c("bill_length_mm", "bill_depth_mm", 
                                  "flipper_length_mm", "body_mass_g")]

# apply mean to these columns
apply(penguins_numeric, 2, mean, na.rm = TRUE)  

```

Note the `2` stays for `MARGIN = 2` (apply on the columns). If you write 1 it means rows. 

Type `?apply` in your Console for more info.

---

## Option 3: Replacing for loops with apply() functions

The `apply()` family refers to several related functions you can use:

| Function   | Use Case                                          | Input Type                    | Output Type       | Why Use It                                      |
|------------|---------------------------------------------------|-------------------------------|-------------------|-------------------------------------------------|
| `apply()`  | Apply a function across **rows or columns**       | Matrix, data frame            | Vector or matrix  | For row/column-wise operations                  |
| `lapply()` | Apply a function to **each element**              | List, data frame              | List              | Keeps outputs as list (safe, flexible)          |
| `sapply()` | Like `lapply()` but **simplifies** output         | List, data frame              | Vector or matrix  | Shorter output, easier to read                  |
| `vapply()` | Like `sapply()`, but **requires output type**     | List, data frame              | Vector or matrix  | Safer than `sapply()` (explicit type checking)  |
| `tapply()` | Apply a function to **vector values by group**    | Vector + Factor(s)            | Vector            | For group-wise summaries (e.g., mean by sex) |

<br>

You don’t need to memorize each one — but remember there are multiple apply() options. Look them up and pick the one that fits your task!

</br>

---

## Option 3: Replacing for loops with apply() functions

```{r echo = TRUE, results = "hide"}

# Use numeric columns from penguins dataset
num_data <- penguins[ , 3:6]

# apply(): function across columns (MARGIN = 2)
apply(num_data, 2, mean, na.rm = TRUE)

# lapply(): list output
lapply(num_data, mean, na.rm = TRUE)

# sapply(): simplified vector output
sapply(num_data, mean, na.rm = TRUE)

# vapply(): safe version of sapply()
vapply(num_data, mean, na.rm = TRUE, FUN.VALUE = numeric(1))

```

::: callout-tip

#### Tips for the `apply()` Family

- `apply()` works best with **matrix-like** structures  
- `lapply()` is **list-safe**, but `sapply()` is more **concise** 
- `vapply()` avoids surprises by **enforcing output type** (here returns a single numeric value)

:::


# 3. While Loops {.slide .center .middle}

---

## Definition of While Loops

We focused on "for loops" because they are the most common, but R, and many other programming languages, also supports "while loops"

**How while loops works:**

- Evaluate a condition
- If the condition is `TRUE`, run the loop body
- Start over: re-evaluate the condition
- Repeat until the condition is `FALSE` (then the while loop is over)

---

## While Loop Syntax

Syntax:
```
while (condition to be evaluated) {
  statement(s)
}
```

Example:
```r
counter <- 1

while(counter <= 4) {
  print(counter)
  counter <- counter + 1
}
```

<!-- 
set the counter outside the loop, usually to 1
set a condition to be evaluated: here the condition says the counter has to be smaller or equal to 4
if the condition is TRUE, the loop is executed, here we only print(counter)
thus the first time it prints 1
but if we leave it as it is (without the last line of code), the while loops will keep going infinitely: we need a way to break the loop
thus we increment our counter inside the loop by redefining it as counter + 1
-->
---

## While Loop Examples

Take the previous code, but this time print `counter` also at the end:

```r
counter <- 1

while(counter <= 3) {
  print(counter)
  counter <- counter + 1
  print(counter)
}
```

Why are the results different from the previous code?

---

## While Loop Examples

Take the previous code, but this time we do not increment our `counter` variable.

```r
counter <- 1
while(counter < 3){
  print(counter)
}
```

What is the output of this code?

---

## While Loop Examples

What is the output of this code?

```r
counter <- 1
while(counter < 4){
  print(counter)
  multiply <- counter * 100
  print(multiply)
  counter <- counter + 1
  print(counter)
}
```

---

## While Loop Uses

While Loops are useful **when you don't know in advance how many times to iterate** — you want the loop to continue **until a condition is met**.

#### Example use cases:

- Keep looping until you get **three heads in a row** from random coin flips  
- Keep accepting user input **until you reach a target number of responses**

While loops require a **"count variable"** to be set outside the loop.

While loops are important but less common than for loops especially for the types of tasks we do in this course. For this reason, we don’t cover them in-depth.


# 4. Introduction to Functions {.slide .center .middle}

---

## From using to writing functions

Today we learn the following concepts:

- Using existing functions
- Writing your own function
- The return statement
- stop()

---

## Using existing functions

Functions are all over R! We have been **using functions** from day 1 of this course. For example:

```{r echo = TRUE, results = "hide"}
a <- c(1:5)
a

sum(a)
mean(a)
```

---

## What is a function?

A function: 

* is a block of code that accomplishes a task

* has a name that typically describes the task, like `sum()` or `mean()`

* takes one or more inputs (e.g., arguments), processes it, and returns one or more outputs

---

## Using existing functions

To **use** a function, we need to know its arguments. 

In the console, type `help()` with the function's name in parenthesis to learn more about a function behavior and its arguments:

```r
help(sum)
help(mean)
```
<!-- mean(c(100, 1, 2, 3, 4, 5, 100, trim = 0.1)) -->

Some functions come from base R, like `sum()` or `mean()`

Other functions belong to specific packages, like `summarize()` or `mutate()` 

---

## Writing custom functions

Other than **using** functions created by others, we can also **write our own functions!**

To do that we need the following:

* **name**: something that describes the task that the function performs
* **arguments (input)**: at least one data structure (e.g., vector, etc.)
* **body**: self-contained lines of code that manipulate the input, defining what the function does
* **return (output):** statement that returns the result of the function's calculations

Syntax:
```r
name <- function(arg1, arg2, ...) { 
    result <- body with code that does things using arg1, arg2, etc.
    return(result)
    }
```
<!--
Note: 
* we can assign the function to a named object like any other object
* typically the body is multiple lines with multiple intermediate variables
-->

---

## Writing custom functions

For example, rather than using the built-in `sum()` function, we could **write our own sum function** to calculate the sum of the elements of a numeric vector, like this:

```r
my_sum <- function(vector) {
  total <- 0
  for (i in seq_along(vector)) {
    total <- total + vector[i]
  }
  return(total)
}

x <- c(2,1,5,2)
my_sum(x)
```

Let's unpack this code in Workbench...

---

## Writing custom functions

**Tip!** Write and test the code outside a function to make sure it works, like this:
```{r, echo = TRUE, results = "hide"}
x <- c(2,1,5,2)

total <- 0  # start with zero, we'll add elements to this

for (i in seq_along(x)) {
  total <- total + x[i]  # add the i-th value of x to total
}

total
```

Then wrap the code into a function and test it again with different inputs:

```{r, echo = TRUE, results = "hide"}

my_sum <- function(vector) {
  total <- 0
  for (i in seq_along(vector)) {
    total <- total + vector[i]
  }
  return(total)
}

x <- c(2, 1, 5, 2)
my_sum(vector = x)

y <- c(4,4)
my_sum(y)
```



<!-- LET'S STOP A SECOND AN UNPACK THIS CODE! DO THIS ON WORKBENCH 
SEE code lecture13_functions.Rmd -->

---

## Writing custom functions

To use a function, we need to **call it** with specific values. 

Our custom function `my_sum()` takes one argument, `vector`. When we call the function, this will be a specific vector, such as `x`, `y`, etc. 

```{r, echo = TRUE, results = "hide"}

x <- c(2,1,5,2)
my_sum(vector = x)

y <- c(4,3,7)
my_sum(y)
```

---

## Advantages of writing custom functions

We use the sum example to illustrate how user-written functions work, but typically if there is already a built-in function in R that does what you need, just use it!

**Instead write your own function when...**

* there is no a built-in function 

* you want to reduce repetitive code and chances for mistakes

* you want to reuse code

* you want organize code (e.g., one function imports the data, another cleans the data, etc.)

---

## Tips for writing functions: name

Use **unique names**:

* If your function name matches the name of an existing R function, your function will replace the existing function for your current session (e.g., write `my_sum`, do not re-write `sum`)
* Unless there is a specific reason (e.g. learning purposes, etc.), do not create a new function like `my_sum()` when R already has a built-in `sum()` function you can use

Use **informative names**: 

* The name should suggest what the function does 
* It should be short
* Avoid reserved words like `if`, `else`, `for`, `function`, etc. To see the full list: type `help(reserved)` in the Console

---

## Tips for writing functions: variables scope

Variables defined inside a function are **not available outside it.** Their scope lies within and is limited to the function itself. Check this out:

```{r, echo = TRUE, results = "hide"}

my_sum <- function(vector) {
  total <- 0
  for (i in seq_along(vector)) {
    total <- total + vector[i]
  }
  return(total)
}

x <- c(2,1,5,2)
my_sum(x)
total
```

---

## The return statement(s)

If you do not write a `return` statement, your output will be the last statement in your code:

```{r, echo = TRUE, results = "hide"}

my_sum <- function(vector) {
  total <- 0
  for (i in seq_along(vector)) {
    total <- total + vector[i]
  }
  total
}

x <- c(2,1,5,2)
my_sum(x)

```

This works, but it's better to use `return` statements for code clarity and readability. **Remember that any code after `return` will be ignored!** Let's see a couple of examples...

---

## Write multiple return statements: Example 1

```{r, echo = TRUE, results = "hide"}

describe_number <- function(x) {
  if (x == 0) {
    return("x is zero (special case!)")
  }
  if (x > 0) {
    return("x is positive")
  } else {
    return("x is negative")
  }
}

```

---

## Write multiple return statements: Example 2

```{r, echo = TRUE, results = "hide"}
my_sum <- function(vector) {
  
  # check 1: input must not be missing
  if (missing(vector)) {
    return("Error: you must provide a vector")
  }
  
  # check 2: input must be numeric
  if (!is.numeric(vector)) {
    return("Error: input must be a numeric vector")
  }
  
  # compute the sum
  total <- 0
  for (i in seq_along(vector)) {
    total <- total + vector[i]
  }
  return(total)
}

# examples
my_sum(c(2, 1, 5, 2))
my_sum(c("one", 2, 3))
my_sum()
```

**Questions:** Why we do not write the "else" in this example? Can you do the same with the previous one?

---

## Write one return statement with multiple outputs

Sometimes you want to return multiple objects and collect them into a list or a vector:

```r
my_sum <- function(vector) {
  total <- 0
  for (i in seq_along(vector)) {
    total <- total + vector[i]
  }
  return(list(total, i))
}

x <- c(2,1,5,2)
y <- my_sum(x)

```

**Questions:** Can we return a vector rather than list here? How? Why a list would generally be a better choice when your function returns multiple objects?

<!-- stop here change to vector with c() and return different elements with y[1] and y[2] change back to list and return different elements again -->

---

## Using stop() when writing a function

Define a function `celsius_to_fahr()` that converts temperatures from Celsius to Fahrenheit using the formula `fahrenheit = (celsius * 9/5) + 32`

```r
celsius_to_fahr <- function(celsius) {
  fahr <- (celsius * 9/5) + 32
  return(fahr)
}

celsius_to_fahr(0)
celsius_to_fahr(-20)
```

---

## Using stop() when writing a function

To ensure this function works properly, the argument must be numeric. Otherwise, the conversion between the two temperature scales will not work.

Solution: use `if-else` to verify if the provided argument is numeric; if not, use `stop()` to raise an error and stop the function execution

```r
celsius_to_fahr <- function(celsius) {
  if (!is.numeric(celsius)) {
    stop("error: input must be a number, try again")
  } else {
    fahr <- (celsius * 9/5) + 32
    return(fahr) 
  }
}

celsius_to_fahr("zero")
```

**Questions:** What is the difference between stop and return? What happens if you rewrite Example 2 (slide 45) using stop instead than return? 

<!-- 
`stop():` interrupts the execution of the function and throw an error message when a condition is not met 

`print()`: doesnt stop the function execution just print the message
so the first enforces a condition that must be met for the function to execute, and prints a message (which is optional); the other is only a message
-->

---

## 💻 Practice writing functions 

### Download today's class materials from the website. 

---

## To print these slides as pdf

Click on the icon bottom-right corner \> Tools \> PDF Export Mode \> Print as a Pdf
