---
title: "Computing for the Social Sciences: Lecture 17"
subtitle: "Topics: Reproducible Workflow"
author: "Sabrina Nardin, Fall 2025"
format:
  revealjs:
    theme: simple
    slide-number: true
    incremental: false
    css: ../style/styles.css
    code-overflow: wrap
    highlight-style: github
    ratio: 16:9
    chalkboard: true
    toc: false
    center: true
---

# Agenda {.center}

::: {.agenda-list}

1. Exam 2 Overview + Final Project Check-in

2. Tips to improve your R Workflow

3. Deep-dive into R Markdown

4. Reflections on AI and Programming
:::

<span style="font-size: 0.8em; color: #666;">
*Slides last updated on **December 1, 2025**. Slides authored by Sabrina Nardin. AI used to polish slides style and fix typos.*
</span>

# 1. Exam 2 Overview + Final Project Check-in {.slide .center .middle}

---

## Exam Format

### Identical to Exam 1 or very similar. Multiple-choice questions and short open-ended questions, such as:

- *Classical multiple choice* — select the correct answer from several options  
- *Spot the typo* — identify and fix syntax or logic errors
- *Matching questions* — pair code snippets with their descriptions or outputs  
- *Debugging questions* — code runs but gives unexpected results; explain *why*  
- *Predict the output* — given a short R snippet determine what the code will return
- *Open-ended questions or write code questions* — brief written explanations and interpretations of results; write code (possible topics: conditionals, loops, functions, base R, data structures)  

---

## Exam Logistics

- Topics covered: Lectures 9-17 (inclusive)

- Time: 60 minutes

- It is closed-book, but you may bring pen and paper for scratch work. See Syllabus for details

- Arrive on time: If you arrive late, you can still take the exam, but no additional time can be given

- Make-up and absence policy: See Syllabus

---

## How to Prepare for the Exam

1. **Review class materials:** slides, in-class code, and assigned readings  

2. **Review previous homework (especially 3 and 4)** re-run your code and check you understand each step  

3. **Review the in-class prep quizzes we have done** exam questions will be similar 

4. **Use AI strategically:** once you’ve done all the above, ask AI to generate additional practice questions of different types and level of difficulty. Starting with AI before reviewing materials can backfires

::: {.callout-tip}
Preparation works best when you write and run code yourself, review your mistakes, and make sure you can explain *why* each answer is correct.
:::

---

## Final Project Check-In

### When: 
Wednesday, December 3, 2025 after the exam (~2:30–3:20 PM)

### What: 
Optional 1:1 meetings (~10 minutes) with the instructor or TA to discuss your final project. Please come with whatever you’ve completed so far and any questions you’d like to go over.

### Sign up:
First come, first served. If many students plan to attend, a sign-up link will be provided.


# 2. Tips to improve your R Workflow {.slide .center .middle}


## Tip 1: Begin every session with a clean, empty workspace

When you run an R session, your workspace includes everything created during that session: libraries that you loaded with `library()`, and several user-created objects (objects, data, functions, etc.)

**Goal: Don’t preserve the workspace. Preserve the code that creates the workspace.** The next slides explain how and why! 

---

## Tip 1: Begin every session with a clean, empty workspace

### How?

When you close a session, by default R will prompt you to save your workspace with the question *“Do you want to save your workspace?”* If you choose to save it, R will reload that same workspace the next time you open it. **Instead, change the default settings:**
  
  * "Do you want to save your workspace?" No
  
  * Tools > Global Options > General 
      * uncheck “Restore .Rdata into workspace at startup”
      * set "Save workspace to .RData on exit" to Never

---

## Tip 1: Begin every session with a clean, empty workspace

### Why? 

**Reproducibility:**  
Saving everything in the code allows others to reproduce your work and enables you to re-run the code to recreate the workspace.

**Fewer Mistakes:**  
Saved workspaces can hide objects you forgot about or didn’t create in this script. For example:
  - Objects created in other scripts may still be lurking
  - You may have loaded packages or data you don’t remember
  - Objects may have been modified without updating the code

---

## Tip 2: Restart R often (Session → Restart R)

Restart R when you want to ensure your script runs from a fresh session:

- More reliable than `rm(list = ls())` or the broom icon
(those remove objects but do not unload packages you loaded earlier)

- Restarting keeps RStudio open, but clears everything underneath, giving you a clean workspace

- Useful for debugging when other fixes don’t work and the code looks correct

---

## Tip 3: Use a Project-based workflow

In this class, we have been using a project-based workflow for everything **by creating a project file `.Rproj` in every repo!** (Remember to create one for your final project, see [Lecture 7](https://brinasab.github.io/css-website/lectures/07-slides.html#/easier-solution-use-rstudio-projects))

**Value of this approach:**

* keeps materials organized

* helps managing working directories (the working directory is the folder that R takes as *default directory* every time you try to access files, scripts, etc.)

---

## Article "Good enough practices in scientific computing"

### Useful Article with more tips to improve your workflow: 
<https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1005510>


# 3. Deep-dive into R Markdown {.slide .center .middle}

---

## R Markdown Recap

**R Markdown is a file format (with extension `.Rmd`) that:**

- combines R code, using code chunks, and text in a single document
- makes it easy to create reproducible reports, analyses, and presentations
- you might hear about Quarto (with extension `.qmd`): a newer alternative to R Markdown (if you learn R Markdown, you can quickly adjust to Quarto)

---

## R Markdown Recap

### Three Main Components

1.  **YAML header** for document title, author, output format, etc. Insert it at the top between `---`

2.  **Text** regular Markdown for headers, lists, links, etc.

3.  **Code chunks** for code. Insert it with triple backticks ```` ``` ````

::: {.fragment .setup-box .callout-tip title="Code Chunk Tips"}
[<strong>Three ways to insert a code chunk in your R Markdown:</strong>]{style="font-size: 1.2em"}

-   Use the RStudio toolbar\
-   Type manually ```` ```{r} ```` and ```` ``` ````
-   Use a shortcut
    -   Mac: ⌘ + ⌥ + I\
    -   Windows: Ctrl + Alt + I
:::

---

## R Markdown vs R Script

### When to use a script vs. a Rmd

- When you want to experiment, test functions, debug code

- When you do not need to communicate output or results to others

- For scripts that need to run top-to-bottom without narrative

- For larger projects where you could have some code in R scripts (e.g., data cleaning, analysis pipelines, tests, etc.), and others (e.g., the final-narrated versions) in a Rmd file

---

## R Markdown: What We Know

* Inserting a YAML Header with basic elements
* Formatting text (italics, bold, etc.)
* Formatting titles 
* Inserting lists (ordered and unordered)
* Inserting images, tables, and hyperlinks
* Rendering a Rmd into pdf, html, md, etc.

---

## R Markdown: What We Learn Today

* How the knitting process works
* Inline code
* Code chunks names
* Code chunks options
* Global options
* New YAML header elements

---

## R Markdown: The Knitting Process

When you "knit" your document, the following sequential things happen, as [illustarated here](https://bookdown.org/yihui/rmarkdown-cookbook/rmarkdown-process.html)

1.  R Markdown sends the `.Rmd` file to `knitr` http://yihui.name/knitr/

2.  Knitr executes all of the code chunks and creates a new plain Markdown `.md` file which includes the code and its output

3.  This plain Markdown file is then converted by `pandoc` into any number of output types including html, PDF, Word document, etc.: http://pandoc.org/

---

## R Markdown: Inline Code

````{verbatim, echo = TRUE}
---
title: "Gun deaths"
author: "Sabrina Nardin"
output: html_document
---

```{r setup, include = FALSE}
library(tidyverse)
library(rcis)
```

## Gun deaths by age

The `gun_deaths` dataset includes data for `r nrow(gun_deaths)` individuals killed by guns.

```{r youth, echo = FALSE}
youth <- gun_deaths %>% filter(age <= 65)
```

We have `r nrow(youth)` individuals who are 65 or younger. 
Only `r nrow(gun_deaths) - nrow(youth)` are older than 65.

````

---

## R Markdown: Inline Code

```{r youth, include = FALSE}
library(tidyverse)
library(rcis)
youth <- gun_deaths %>% filter(age <= 65)
```

The `gun_deaths` dataset includes data for `r nrow(gun_deaths)` individuals killed by guns.

We have `r nrow(youth)` individuals who are 65 or younger. 
Only `r nrow(gun_deaths) - nrow(youth)` are older than 65.

---

## R Markdown: Code Chunks Names

A named code chunk:
````{verbatim, lang = 'markdown', echo = TRUE}
```{r youth-distribution}
# code goes here
```
````

Advantages:

1. Easier to navigate to specific chunks (using the drop-down code navigator in the script editor). Note: Two code chunks should not have the same name.

1. Graphics produced by the chunks have useful names

1. Use it with `cache = TRUE` to set up cached chunks to avoid re-performing expensive computations on every run 

::: {.callout-tip}
`cache = TRUE` saves a chunk's output to a file on memory. On future renders, R checks if the code chunk has changed; if not, R skips re-running the chunk and reuses the cached results (see Ch. 27 R for Data Science)
:::


---

## R Markdown: Code Chunk Options {.center}

::: small
| Option | What it does | Default |
|------------------------|------------------------|------------------------|
| `eval = FALSE` | Code is not run, only shown; results do not appear. Useful for code examples or showing code with errors. | `eval = TRUE` |
| `include = FALSE` | Code is run, but both code and results are hidden. Useful for setup code you don't want to display. | `include = TRUE` |
| `echo = FALSE` | Code is run, results are shown, but code is hidden. Good when readers only need the output. | `echo = TRUE` |
| `error = TRUE` | Code is run and shown, even if it throws an error. Useful when demonstrating broken code. | `error = FALSE` |
| `message = FALSE` / `warning = FALSE` | Code is run, but messages and warnings are hidden from the output. | `message = TRUE` / `warning = TRUE` |
:::

Check [this table](https://r4ds.had.co.nz/r-markdown.html#chunk-options) and [options](http://yihui.name/knitr/options/) for more.

---

## R Markdown: Global Options

Use global chunk options when you want the same settings to apply to ALL code chunks in your R Markdown file.

Global options are defined in a code chunk at the top of your Rmd file with `knitr::opts_chunk$set()` 

For example, if you want to hide the code by default:

```r
knitr::opts_chunk$set(
  echo = FALSE    # shows results but hide codes 
)
```
---

## R Markdown: New YAML Header Elements

````{verbatim, echo = TRUE}
---
title: "Gun deaths"
author: "Sabrina Nardin"
output: html_document
---
````

---

## YAML Header: Table of Contents

````{verbatim, echo = TRUE}
---
title: "Gun deaths"
author: "Sabrina Nardin"
output: 
  html_document
    toc: true
    toc_depth: 2
---
````

---

## YAML Header: Apparence and Style

````{verbatim, echo = TRUE}
---
title: "Gun deaths"
author: "Sabrina Nardin"
output: 
  html_document
    theme: readable
    fig_width: 8
    fig_height: 6
---
````

---

## YAML Header: Date as string

We can add a date as a simple string, and manually modify it every time we modify our file:

````{verbatim, echo = TRUE}
---
title: "Gun deaths"
output: html_document
date: "December 1, 2025"
---
````

But we can also add it automatically!

---

## YAML Header: Date with lubridate 

````{verbatim, echo = TRUE}
---
title: "Gun deaths"
output: html_document
date: "`r lubridate::today()`"
---
````

For more on the `lubridate` package, see Chapter 16 of R for Data Science

---

## YAML Header: Date with Sys.Date

````{verbatim, echo = TRUE}
---
title: "Gun deaths"
output: html_document
date: "`r Sys.Date()`"
---
````

`Sys.Date` returns the current day in the current time zone. The default uses the format year/month/day.

---

## YAML Header: Date with Sys.time

````{verbatim, echo = TRUE}
---
title: "Gun deaths"
output: html_document
date: "`r format (Sys.time(), '%d, %B, %Y')`"
---
````

`Sys.time` returns an absolute date-time value which can be converted to various time zones, and it more easily is customizable. This format says we want the date as: numeric day, full name of the month, and full (4-digit) year, but it can be changed

---

## 💻 Practice with Rmd

Download today's class materials and complete the tutorial on Rmd

---

## R Markdown More Info and Tips

For the final project, experiment with the R Markdown options we have seen today, in addition to the things you already know. 

For more, see [R for Data Science 1st Edition](http://r4ds.had.co.nz):

- Chapter 27 "R Markdown"
- Chapter 28 "Graphics for communication"
- Chapter 29 "R Markdown formats" 


# 4. Reflections on AI and Programming {.slide .center .middle}

---

## Reflecting on AI Use for Programming

One part of my job is teaching you R fundamentals. Another (new) part is helping you learn how to use AI as a learning tool, as I think AI is here to stay.

I'd like to do an activity on reflecting on your AI use in this course using this [Google Doc](https://docs.google.com/document/d/1RyzfQ022OdChYVKWW6VPhEnJgjUmdyYmhtumHld-cso/edit?usp=sharing)

