---
title: "Computing for the Social Sciences: Lecture 18"
subtitle: "Topics: Control Flow with Loops (cont.); Functions"
author: "Sabrina Nardin, Summer 2026"
format:
  revealjs:
    theme: simple
    slide-number: true
    incremental: false
    css: ../style/styles.css
    code-overflow: wrap
    highlight-style: github
    ratio: 16:9
    chalkboard: true
    toc: false
    center: true
---


# Agenda {.center}

::: {.agenda-list}
1. Alternatives to For Loops in R
2. While Loops
3. Introduction to Functions
:::

<span style="font-size: 0.8em; color: #666;">
*Slides last updated on **`r format(Sys.Date(), '%B %d, %Y')`**. Slides authored by Sabrina Nardin. AI used to polish slides style and fix typos.*
</span>

<!--

---

### Slides are in progress and will be updated before class.

::: {.content-hidden}
-->


# 1. Alternatives to For Loops in R {.slide .center .middle}

---

## Last Lecture: we used a for loop with a dataframe

In [lecture 16](https://brinasab.github.io/css-website-summer/lectures/16-slides.html#/using-for-loops-with-data-frames) we learned how to apply a for loop to calculate mean value of several columns in a dataframe. Like this:

```{r echo = TRUE, results = "hide"}
library(tidyverse)
library(palmerpenguins)

output <- vector(mode = "double", length = ncol(penguins))

for (i in seq_along(penguins)) {
  #print(i)
  #print(penguins[i])
  output[i] <- mean(penguins[[i]], na.rm = TRUE)
}

output
```

Let's examine this code...

- `i` will index each column
- `output[i]` saves the results in the output vector at the correct position
- `penguins[[i]]` ` pulls out the column's values as a plain vector 

---

## Last Lecture we ended with this slide

For loops make iteration explicit and are an important programming tool. You will encounter them in many programming languages.

**But R is a VECTORIZED language, built to operate on a whole dataframe column at once rather than element by element.** 

That means it has functions that can do the *same job* as a for loop, just in a way that fits R's strengths better. Next week we'll explore them:

- `across()` (dplyr): apply a function to many columns in one line
- `map()` (purrr): apply a function to every element of a list or column
- `apply()` family (base R): the classic vectorized alternative to for loops

---

## Why We Even Learn Loops in R?

Many tasks in R can and should be done **without writing a for loop**. 

So why do we even learn loops in R? Because understanding loops is important:

- to build logic and control flow skills
- for non-vectorized tasks**, like row-by-row operations
- to build more complex workflows
- to be able to read and understand other people code
- for general literacy on programming

> Even if you don’t use loops every day, knowing how they work will strengthen your skills as coder.

<!--

SO WHY A FOR LOOP ?? 

The honest answer for a dataframe-column context is closer to:

1. You'll read loops in other people's/older code, even if you rarely write them yourself for this kind of task — that's a literacy argument, not a capability one.

2. across() only fits when the tidy-select + one-function pattern fits. The moment the task per column is a genuinely custom, multi-step, branching procedure (rather than "apply this one function"), a loop becomes clearer than contorting it into across().

3. The real can't-do-it-with-across() cases tend to live outside single-dataframe, column-independent tasks — R4DS's own examples for "why iterate" are things like reading in many files, fitting many models and saving each one, or writing out many objects — tasks with side effects, not a column transformation.

Given that, I'd suggest replacing the running-total-over-columns slide with something more defensible — want me to draft a slide built around one of R4DS's actual examples (e.g., looping to read/fit/save several things), or would you rather just soften the claim to "loops are the general tool and you'll need to recognize them in the wild," without insisting on a dataframe-specific counterexample?

ALso add example or looping over rows

data <- data.frame(
  name = c("Alice", "Bob", "Carlos"),
  score = c(85, 92, 78)
)

for (i in seq_len(nrow(data))) {
  print(data[i, ])
}

-->

---


## Three Main Alternatives to For Loops in R

**When working with dataframes**, R provides the following alternatives to for loop:

| Function Family        | Package     | Best Used For                                  | Docs |
|------------------------|-------------|------------------------------------------------|-------------|
| `map_*()`              | purrr     | Applying functions to elements of lists or columns | [map docs](https://purrr.tidyverse.org/reference/map.html) |
| `across()`             | dplyr     | Applying the same function to multiple columns inside `mutate()` or `summarize()` | [across docs](https://dplyr.tidyverse.org/reference/across.html) |
| `*apply()`             | base R      | Row/column-wise operations on matrices or data frames | [apply docs](https://stat.ethz.ch/R-manual/R-devel/library/base/html/apply.html) |

<br>

The asterisk `*` is a placeholder for a family of related functions (e.g., `map_dbl()`, `map_chr()`, `sapply()`, etc.)

</br>

<!--
1. `map_*()` functions from `purr` <https://purrr.tidyverse.org/reference/map.html>
2. `across()` function from `dplyr` <https://dplyr.tidyverse.org/reference/across.html>
3. `*apply()` functions from base R <https://stat.ethz.ch/R-manual/R-devel/library/base/html/apply.html>

The asterisk `*` is a placeholder indicating a *family* of related functions (e.g., `map_dbl()`, `map_chr()`, `sapply()`, etc.).

- Use the `map_*()` functions from the *purrr* package for flexible iteration  
- Use `across()` with `summarize()` or `mutate()` for column-wise operations  
- Use `apply()` to iterate over rows or columns in a matrix or data frame
-->

---

## Option 1: Replacing for loops with map() functions

<!-- INTRO: 
As we have seen in the slides and practice, we have to write a lot of code for a for loop, for example to calculate a straightforward operation like the mean or median etc. so R gives us a shortcut: map functions which come from the purrr package in R and are much more compact (go to the documentation page)
-->

**The first option to replace a for loop is the family of `map_*()` functions from the `purrr` package.**

**What it does:** applies a function to each element of a list or vector (for a data frame: each column), and returns the results in the type of output you choose

**Best use:** flexible, all-purpose iteration; unlike `across()`, it does not need to live inside a `dplyr` verb like `summarize()` or `mutate()`


---

## Option 1: Replacing for loops with map() functions


The so-called "map functions" come from the `purr` package in R: <https://purrr.tidyverse.org/reference/map.html>

**There are several `map_*()` functions**  each creates a different type of output:

- `map()` makes a list
- `map_lgl()` makes a logical vector
- `map_int()` makes an integer vector
- `map_dbl()` makes a double vector
- `map_chr()` makes a character vector

This is the same idea as in the `for loop` when we specify the `mode` of our output vector.

Let's see a few examples using the `penguins` data.

---

## Option 1: Replacing for loops with map() functions

To use these properly, you need to pick the appropriate `map()` function (see previous slide) and specify at least two main arguments:

- what you are iterating over
- what you are calculating

Here we are calculating the mean of all numeric columns of the penguins dataframe:
```{r echo = TRUE, results = "hide"}

penguins |> 
  select(where(is.numeric)) |>
  map_dbl(mean, na.rm = TRUE)

```

what does `select(where(is.numeric))` do? And why we picked `map_dbl()` here?

<!-- 
answers:
- we are selecting only numeric columns
- expect mean to have numbers afer comma
also note:
everything easier than a loop to both write and understand 
operates well with dplyr -->

---

## 💻 Option 1 Practice: using functions from the map family

Copy and paste this code (from the previous slide) into R and run it:

```r
library(tidyverse)
library(palmerpenguins)

penguins |> 
  select(where(is.numeric)) |>
  map_dbl(mean, na.rm = TRUE)
```

**Your tasks:**

1. Modify the code to compute the *median* (instead of the mean) of each numeric column
2. In the original code, replace `map_dbl()` with `map()`. What changes in the output?


<!--
```
# 1. median of each numeric column
penguins |> 
  select(where(is.numeric)) |>
  map_dbl(median, na.rm = TRUE)
  
# 2. map() returns a list, map_dbl() returns a named double vector
penguins |> 
  select(where(is.numeric)) |>
  map(mean, na.rm = TRUE)


SKIPPED THIS 

3. Challenge: Use a map function to count the *number of missing values (`NA`) every column* of `penguins`. Hint: `sum(is.na(.x))` counts the `NA`s in a column; which `map_*()` variant returns integers?
  
# 3. number of NAs in every column (integers, so map_int)
penguins |> 
  map_int(~ sum(is.na(.x)))

# or "col" is just a name we choose: it stands for the current column
penguins |> 
  map_int(function(col) sum(is.na(col)))
  
"Count the NAs" is not one existing function — it's two chained together: is.na() then sum()
Since there's no built-in function to name, you write the recipe yourself
In the shorthand: ~ means "here comes a mini-function," and .x is the placeholder for the current column (it plays the same role as col in the explicit version)
  
```
-->

---

## Option 2: Replacing for loops with across()

**Another option to replace a for loop is the `across()` function from `dplyr`**

**What it does:** applies the same operation (e.g., mean) to multiple columns at once  

**Best use:** since it comes from `dplyr`, it works seamlessly inside `dplyr` verbs like `mutate()` and `summarize()` which these are its favorite verbs to work with! You might remember this function from [Lecture 7](https://brinasab.github.io/css-website-summer/lectures/07-slides.html#/across)

---

## Option 2: Replacing for loops with across()

This example calculates the mean of several columns in a data frame. We used it in Lecture 16, when we rewrote it using a for loop:

```{r echo = TRUE, results = "hide"}

penguins |> summarize(avg_bill_length = mean(bill_length_mm, na.rm = TRUE))
penguins |> summarize(avg_bill_depth = mean(bill_depth_mm, na.rm = TRUE))
penguins |> summarize(avg_flipper_length = mean(flipper_length_mm, na.rm = TRUE))
penguins |> summarize(avg_body_mass = mean(body_mass_g, na.rm = TRUE))
                       
```

Now, we rewrite it again but this time using `across()`

---

## Option 2: Replacing for loops with across()

Rewrite previous example using `across()`:

```{r echo = TRUE, results = "hide"}

penguins |> 
  summarize(
    across(
      .cols = where(is.numeric),           # select columns by type
      .fns = ~ mean(.x, na.rm = TRUE)))    # apply mean, skipping NA
```

<br>

**The function `across()` has two main arguments:**

- **`.cols`**: the columns to operate on. You can select them by position, name, or type - in this example, by type using `where(is.numeric)`

- **`.fns`**: the function, or list of functions, to apply to each column (in this example, we use `mean()` and `.x` is a placeholder for the current column being processed)

You can omit `.cols` and `.fns` if you pass the arguments in the correct order, like that:
`penguins |> summarize(across(where(is.numeric), ~ mean(.x, na.rm = TRUE)))`

<!--
TEACHING INSIGHT

penguins |> 
  summarize(across(where(is.numeric), mean)) |>   # works! (but NAs give NA results)
  print(width = Inf)
  
"In across(), the moment your function needs any extra argument (like na.rm = TRUE), you must wrap it: ~ mean(.x, na.rm = TRUE). Passing the argument loose after the function name is the old way and is now deprecated. In map(), both styles still work
-->

</br>

---

## Option 2: Replacing for loops with across()

*Tip:* you can also add names to the newly computed means to improve readabilty

```{r echo = TRUE, results = "hide"}

penguins |> 
  summarize(
    across(
      .cols = where(is.numeric),                   
      .fns = ~ mean(.x, na.rm = TRUE),             
      .names = "avg_{.col}"                   
    )
  )
```

<br>

**What does `\"avg_{.col}\"` mean?**

- `"avg_"` is a literal prefix that will be added to each new column name.  
- `{.col}` is a placeholder that will be replaced with the original column name.

So if you're applying `mean()` to a column named `bill_length_mm`, the result will be named `avg_bill_length_mm`

</br>

---

## 💻 Option 2 Practice: using across

Copy and paste this code (from the previous slide) into R and run it:

```r
penguins |> 
  summarize(
    across(
      .cols = where(is.numeric),                   
      .fns = ~ mean(.x, na.rm = TRUE),             
      .names = "avg_{.col}"                   
    )
  )
```

**Your tasks:**

1. Compute these means **separately for each species**. Hint: add one `dplyr` verb before `summarize()`
2. Change the code to compute the **median**, and rename the new columns `"med_{.col}"`
3. Bonus: apply the mean only to the columns whose name **ends with `"_mm"`**. Hint: Select columns by name patterns instead of type with `ends_with()` in `.cols`

<!--  SOLUTIONS 
# 1. and 2. median by species, with new names
penguins |> 
  group_by(species) |>
  summarize(
    across(
      .cols = where(is.numeric),
      .fns = ~ median(.x, na.rm = TRUE),
      .names = "med_{.col}"
    )
  )

# 3. select columns by name pattern instead of type
penguins |> 
  summarize(
    across(
      .cols = ends_with("_mm"),
      .fns = ~ mean(.x, na.rm = TRUE),
      .names = "avg_{.col}"
    )
  )


FOR FUTURE: add one practice with across and filter + if_any()

-->

---

## Option 3: Replacing for loops with apply() functions

**Finally, a third option to replace a for loop in R is using one of the `apply()` functions from base R**

**What it does:** applies the same operation (e.g., mean) across the rows or the columns of a data frame

**Best use:** when working outside tidyverse verbs (this is from base R!); applying a function across columns is the most common use

---

## Option 3: Replacing for loops with apply() functions

We looked at this example earlier (which calculates the mean of several columns in a data frame) and rewrote it using a `for` loop. Now we rewrite it again but this time using `apply()`:

```{r echo = TRUE, results = "hide"}

penguins |> summarize(avg_bill_length = mean(bill_length_mm, na.rm = TRUE))
penguins |> summarize(avg_bill_depth = mean(bill_depth_mm, na.rm = TRUE))
penguins |> summarize(avg_flipper_length = mean(flipper_length_mm, na.rm = TRUE))
penguins |> summarize(avg_body_mass = mean(body_mass_g, na.rm = TRUE))
                       
```

---

## Option 3: Replacing for loops with apply() functions

Rewrite the previous example using `apply()`:

```{r echo = TRUE, results = "hide"}

# manually select only numeric column using base R syntax
penguins_numeric <- penguins[ , c("bill_length_mm", "bill_depth_mm", 
                                  "flipper_length_mm", "body_mass_g")]

# apply mean to these columns
apply(penguins_numeric, 2, mean, na.rm = TRUE)  

```

Note the `2` stays for `MARGIN = 2` (apply on the columns). If you write 1 it means rows. 

Type `?apply` in your Console for more info.

---

## Option 3: Replacing for loops with apply() functions

The `apply()` family refers to several related functions you can use:

| Function   | Use Case                                          | Input Type                    | Output Type       | Why Use It                                      |
|------------|---------------------------------------------------|-------------------------------|-------------------|-------------------------------------------------|
| `apply()`  | Apply a function across **rows or columns**       | Matrix, data frame            | Vector or matrix  | For row/column-wise operations                  |
| `lapply()` | Apply a function to **each element**              | List, data frame              | List              | Keeps outputs as list (safe, flexible)          |
| `sapply()` | Like `lapply()` but **simplifies** output         | List, data frame              | Vector or matrix  | Shorter output, easier to read                  |
| `vapply()` | Like `sapply()`, but **requires output type**     | List, data frame              | Vector or matrix  | Safer than `sapply()` (explicit type checking)  |
| `tapply()` | Apply a function to **vector values by group**    | Vector + Factor(s)            | Vector            | For group-wise summaries (e.g., mean by sex) |

<br>

You don’t need to memorize each one — but remember there are multiple `apply()` options (like there are multiple `map()` options). Look them up and pick the one that fits your task!

</br>

---

## Option 3: Replacing for loops with apply() functions

```{r echo = TRUE, results = "hide"}

# Use numeric columns from penguins dataset
num_data <- penguins[ , 3:6]

# apply(): function across columns (MARGIN = 2)
apply(num_data, 2, mean, na.rm = TRUE)

# lapply(): list output
lapply(num_data, mean, na.rm = TRUE)

# sapply(): simplified vector output
sapply(num_data, mean, na.rm = TRUE)

# vapply(): safe version of sapply()
vapply(num_data, mean, na.rm = TRUE, FUN.VALUE = numeric(1))

```

::: callout-tip

#### Tips for the `apply()` Family

- `apply()` works best with **matrix-like** structures  
- `lapply()` is **list-safe**, but `sapply()` is more **concise** 
- `vapply()` avoids surprises (e.g. unpredictable output type) by **enforcing output type** (here returns a single numeric value)

:::

<!--
The "surprises" are sapply()'s unpredictable output type. 
sapply() tries to guess how to simplify its result, and what you get back depends on the data:

If every element returns one value → you get a vector
If every element returns the same number of values (>1) → you get a matrix
If elements return different lengths → simplification fails silently and you get a list (same as lapply())
-->


---

## 💻 Option 3 Practice: using functions from the apply family 

Copy and paste this code into your script and run it:

```r
num_data <- penguins[ , 3:6]
sapply(num_data, median, na.rm = TRUE)
```

**Your tasks:**

1. Rewrite the code above using `lapply()`. How is the output different?
2. A common pitfall! Run `apply()` on the **full** dataset: `apply(penguins, 2, mean, na.rm = TRUE)`. What do you get, and why? Hint: `apply()` first converts its input to a **matrix**, and a matrix can hold only **one** data type. What happens when a data frame mixes numbers and text?

<!-- SOLUTIONS

```r
# 1. lapply() returns a list (one element per column),
#    while sapply() simplifies the result to a named vector
lapply(num_data, median, na.rm = TRUE)

# 2. every column comes back NA (with warnings!)
apply(penguins, 2, mean, na.rm = TRUE)
```

**Why task 2 fails:** `penguins` mixes numeric and non-numeric columns (e.g., `species`, `island`). When `apply()` converts it to a matrix, *everything* is coerced to character, and `mean()` of a character vector returns `NA` with a warning.

**Takeaway:** with data frames of mixed types, select the numeric columns first — or use `sapply()`/`lapply()`, which work column by column without converting to a matrix.
-->


# 2. While Loops {.slide .center .middle}

---

## Definition of While Loops

We focused on "for loops" because they are the most common, but R, and many other programming languages, also supports "while loops"

**How while loops work:**

- Evaluate a condition
- If the condition is `TRUE`, run the loop body
- Start over: re-evaluate the condition
- Repeat until the condition turns into `FALSE` (then the while loop is over)

---

## While Loop Syntax

Syntax:
```
while (condition to be evaluated) {
  statement(s)
}
```

Example:
```r
counter <- 1

while(counter <= 4) {
  print(counter)
  counter <- counter + 1
}
```

<!--  TEACHING NOTES
set the counter outside the loop, usually to 1
set a condition to be evaluated: here the condition says the counter has to be smaller or equal to 4
if the condition is TRUE, the loop is executed, here we only print(counter)
thus the first time it prints 1
but if we leave it as it is (without the last line of code), the while loops will keep going infinitely: we need a way to break the loop
thus we increment our counter inside the loop by redefining it as counter + 1
-->

---

## While Loop Examples

Take the previous code, but this time print `counter` also at the end:

```r
counter <- 1

while(counter <= 3) {
  print(counter)
  counter <- counter + 1
  print(counter)
}
```

Why are the results different from the previous code?

---

## While Loop Examples

Take the previous code, but this time we do not increment our `counter` variable. 
What is the output of this code?


```r
counter <- 1
while(counter < 3){
  print(counter)
}
```

---

## While Loop Examples

Observe this last one. What is the output of this code?

```r
counter <- 1
while(counter < 4){
  print(counter)
  multiply <- counter * 100
  print(multiply)
  counter <- counter + 1
  print(counter)
}
```

---

## While Loop Uses

While Loops are useful **when you don't know in advance how many times to iterate:** you want the loop to continue **until a condition is met.**

#### Example use cases:

- Keep looping until you get three heads in a row from random coin flips  
- Keep accepting user input until you reach a target number of responses

While loops typically require a **counter variable**: set it *outside* the loop, and update it *inside* the loop, otherwise the loop never ends.

While loops are important but less common than for loops, especially for the types of tasks we do in this course. 


# 3. Introduction to Functions {.slide .center .middle}

---

## From using to writing functions

Today we learn the following concepts:

- Using existing functions
- Writing your own function
- The return statement
- stop()

---

## Using existing functions

Functions are all over R! We have been **using functions** from day 1 of this course. For example:

```{r echo = TRUE, results = "hide"}
a <- c(1:5)
a

sum(a)
mean(a)
```

---

## What is a function?

A function: 

* is a block of code that accomplishes a task

* has a name that typically describes the task, like `sum()` or `mean()`

* takes one or more inputs arguments), processes them, and returns one or more outputs

---

## Using existing functions

To **use** a function, we need to know its arguments. 

In the console, type `help()` with the function's name in parenthesis to learn more about a function behavior and its arguments:

```r
help(sum)
help(mean)
```
<!-- mean(c(100, 1, 2, 3, 4, 5, 100, trim = 0.1)) -->

Some functions come from base R, like `sum()` or `mean()`

Other functions belong to specific packages, like `summarize()` or `mutate()` which belong to `dplyr`

---

## Writing custom functions

Other than **using** functions created by others, we can also **write our own functions!**

To do that we need the following:

* **name**: something that describes the task that the function performs
* **arguments (input)**: at least one data structure (e.g., vector, etc.)
* **body**: self-contained lines of code that manipulate the input, defining what the function does
* **return (output):** statement that returns the result of the function's calculations

Syntax:
```r
name <- function(arg1, arg2, ...) { 
    result <- body with code that does things using arg1, arg2, etc.
    return(result)
    }
```
<!--
Note: 
* we can assign the function to a named object like any other object
* typically the body is multiple lines with multiple intermediate variables
-->

---

## Writing custom functions

In R we can also write our own function! 

For example, we could **write our own sum function** to calculate the sum of the elements of a numeric vector, like this:

```r
my_sum <- function(vector) {
  total <- 0
  for (i in seq_along(vector)) {
    total <- total + vector[i]
  }
  return(total)
}

x <- c(2,1,5,2)
my_sum(x)
```

Let's unpack this code...

---

## Writing custom functions

**Tip!** Write and test the code outside a function to make sure it works, like this:
```{r, echo = TRUE, results = "hide"}
x <- c(2,1,5,2)

total <- 0  # start with zero, we'll add elements to this

for (i in seq_along(x)) {
  total <- total + x[i]  # add the i-th value of x to total
}

total
```

Then wrap the code into a function and test it again with different inputs:
```{r, echo = TRUE, results = "hide"}

my_sum <- function(vector) {
  total <- 0
  for (i in seq_along(vector)) {
    total <- total + vector[i]
  }
  return(total)
}

x <- c(2, 1, 5, 2)
my_sum(vector = x)

y <- c(4,4)
my_sum(y)
```


<!-- LET'S STOP A SECOND AN UNPACK THIS CODE! DO THIS ON WORKBENCH. SEE code lecture13_functions.Rmd -->

---

## Writing custom functions

To use a function, we need to **call it** with specific values. 

Our custom function `my_sum()` takes one argument, `vector`. When we call the function, this will be a specific vector, such as `x`, `y`, etc. 

```{r, echo = TRUE, results = "hide"}

x <- c(2,1,5,2)
my_sum(vector = x)

y <- c(4,3,7)
my_sum(y)
```

---

## Advantages of writing custom functions

We wrote a sum function to illustrate how user-written functions work, but typically if there is already a built-in function in R , it's best to use it (vs. rewriting it)

**Instead write your own function when...**

* there is no a built-in function 

* you want to reduce repetitive code and chances for mistakes

* you want to reuse code

* you want organize code (e.g., one function imports the data, another cleans the data, etc.)

---

## Tips for writing functions: name

Use **unique names**:

* If your function name matches the name of an existing R function, your function will replace the existing function for your current session (e.g., write `my_sum`, do not re-write `sum`)
* Unless there is a specific reason (e.g. learning purposes, etc.), do not create a new function like `my_sum()` when R already has a built-in `sum()` function you can use

Use **informative names**: 

* The name should suggest what the function does 
* It should be short
* Avoid reserved words like `if`, `else`, `for`, `function`, etc. To see the full list: type `help(reserved)` in the Console

---

## Tips for writing functions: variables scope

Variables defined inside a function are **not available outside it.** Their scope lies within and is limited to the function itself.

Run this in a fresh R session: `total` exists only *inside* `my_sum()`:

```r
my_sum <- function(vector) {
  total <- 0
  for (i in seq_along(vector)) {
    total <- total + vector[i]
  }
  return(total)
}

x <- c(2,1,5,2)
my_sum(x)
total   # Error: object 'total' not found

```
---

## The return statement(s)

If you do not write a `return` statement, your output will be the last statement in your code:

```{r, echo = TRUE, results = "hide"}

my_sum <- function(vector) {
  total <- 0
  for (i in seq_along(vector)) {
    total <- total + vector[i]
  }
  total
}

x <- c(2,1,5,2)
my_sum(x)

```

This works, but it's best practice to use `return` statements for code clarity and readability. **Remember that any code after `return` will be ignored!** Let's see a couple of examples...

---

## Write multiple return statements: Example 1

```{r, echo = TRUE, results = "hide"}

describe_number <- function(x) {
  if (x == 0) {
    return("x is zero (special case!)")
  }
  if (x > 0) {
    return("x is positive")
  } else {
    return("x is negative")
  }
}

```

---

## Write multiple return statements: Example 2

```{r, echo = TRUE, results = "hide"}
my_sum <- function(vector) {
  
  # check 1: input must not be missing
  if (missing(vector)) {
    return("Error: you must provide a vector")
  }
  
  # check 2: input must be numeric
  if (!is.numeric(vector)) {
    return("Error: input must be a numeric vector")
  }
  
  # compute the sum
  total <- 0
  for (i in seq_along(vector)) {
    total <- total + vector[i]
  }
  return(total)
}

# examples
my_sum(c(2, 1, 5, 2))
my_sum(c("one", 2, 3))
my_sum()
```

**Questions:** Why we do not write the "else" in this example? Can you do the same with the previous one?

---

## Write one return statement with multiple outputs

Sometimes you want to return multiple objects and collect them into a list or a vector:

```r
my_sum <- function(vector) {
  total <- 0
  for (i in seq_along(vector)) {
    total <- total + vector[i]
  }
  return(list(total, i))
}

x <- c(2,1,5,2)
y <- my_sum(x)

```

**Questions:** Can we return a vector rather than list here? How? Why a list would generally be a better choice when your function returns multiple objects?

<!-- stop here change to vector with c() and return different elements with y[1] and y[2] change back to list and return different elements again -->

---

## Using stop() when writing a function

Define a function `celsius_to_fahr()` that converts temperatures from Celsius to Fahrenheit using the formula `fahrenheit = (celsius * 9/5) + 32`

```r
celsius_to_fahr <- function(celsius) {
  fahr <- (celsius * 9/5) + 32
  return(fahr)
}

celsius_to_fahr(0)
celsius_to_fahr(-20)
```

---

## Using stop() when writing a function

To ensure this function works properly, the argument must be numeric. Otherwise, the conversion between the two temperature scales will not work.

Solution: use `if-else` to verify if the provided argument is numeric; if not, use `stop()` to raise an error and stop the function execution

```r
celsius_to_fahr <- function(celsius) {
  if (!is.numeric(celsius)) {
    stop("error: input must be a number, try again")
  } else {
    fahr <- (celsius * 9/5) + 32
    return(fahr) 
  }
}

celsius_to_fahr("zero")
```

**Question:** What is the difference between `stop()` and `return()`? 

<!-- 
`stop():` interrupts the execution of the function and throw an error message when a condition is not met 

`print()`: doesnt stop the function execution just print the message
so the first enforces a condition that must be met for the function to execute, and prints a message (which is optional); the other is only a message
-->

---

## Using stop() when writing a function

Most functions exit in one of two ways: they either `return()` a value, indicating success, or they `stop()` and throw an error, indicating failure. Therefore:

- `return()`: exits the function NORMALLY, handing back a value
- `stop()`: exits the function WITH AN ERROR, so execution halts, and the console shows `Error:`, and no misleading value is returned. 

---

## Using stop() when writing a function: obseve this comparison


**Approach 1: using `return()` for errors** ❌ 

```r
my_sum <- function(vector) {
  
  # input check with return(): the error message is just a regular value
  if (!is.numeric(vector)) {
    return("Error: input must be a numeric vector")
  }
  
  # compute the sum
  total <- 0
  for (i in seq_along(vector)) {
    total <- total + vector[i]
  }
  return(total)
}

result <- my_sum(c("one", 2, 3))  # no error shown!
result                            # only now we see something is off
result + 10                       # fails HERE, with a confusing message

```

## Using stop() when writing a function: obseve this comparison

**Approach 2: using `stop()` for errors** ✅ 

```r
my_sum <- function(vector) {
  
  # input check with stop(): raise a real error
  if (!is.numeric(vector)) {
    stop("input must be a numeric vector")
  }
  
  # compute the sum (identical to above)
  total <- 0
  for (i in seq_along(vector)) {
    total <- total + vector[i]
  }
  return(total)
}

result <- my_sum(c("one", 2, 3))  # Error: input must be a numeric vector, halts immediately
                                  # with OUR message; result is never created
result + 10                       # in a script, this line is never reached
```

<!--
Note the stop() message is just "input must be a numeric vector" — no "Error:" prefix inside it, since R adds that automatically when the error prints. And the final return(total) stays as return() in both versions: the convention isn't "never use return()," it's return for results, stop for problems.Want to be notified when Claude responds?
-->


---

## 💻 Practice writing functions 

### Download today's class materials from the website for more practice on writing functions

---

## Recap: What We Learned Today

- Using alternatives to for loops: `map_*()` (purrr), `across()` (dplyr), and the `*apply()` family (base R)
- How while loops work: check a condition, run the body, repeat until the condition is `FALSE`
- How to write custom functions: name, arguments, body, and return
- Using `return()` and `stop()` to check inputs and raise errors

---

## To print these slides as pdf

Click on the icon bottom-left corner (the three lines) \> Tools \> PDF Export Mode \> Print as a Pdf

:::
