---
title: "Computing for the Social Sciences: Lecture 19"
subtitle: "Topics: Functions (cont.)"
author: "Sabrina Nardin, Summer 2026"
format:
  revealjs:
    theme: simple
    slide-number: true
    incremental: false
    css: ../style/styles.css
    code-overflow: wrap
    highlight-style: github
    ratio: 16:9
    chalkboard: true
    toc: false
    center: true
---

# Agenda {.center}

::: {.agenda-list}
1. Anonymous Functions 
2. Documenting your Function
3. Functions with the tidyverse
:::

<span style="font-size: 0.8em; color: #666;">
*Slides last updated on **`r format(Sys.Date(), '%B %d, %Y')`**. Slides authored by Sabrina Nardin. AI used to polish slides style and fix typos.*
</span>

<!--


### Slides are in progress and will be updated before class.

::: {.content-hidden}

-->

---

So far, we learned the following concepts:

- Using existing functions
- Writing your own function
- Using return()
- Using stop()

In these slides, we add:

- Anonymous functions
- Documenting your function
- Writing functions with the tidyverse


# 1. Anonymous Functions {.slide .center .middle}

---

## Anonymous Functions

Anonymous functions are **functions without a name**

**When to use them:**

* When the function is very short and you need it only once 

* When another function (like those in the `apply()` family) expects another function as input

Whenever you can: write the function explicitly instead!

---

## Anonymous Functions with a number as input

Imagine we have the following function:

```{r echo = TRUE, results = "hide"}
add <- function(x) {
  return(x + 3) }
add(2)
```

Re-write it as an anonymous function. Note: one line, no name, and call it with `()`:
```{r echo = TRUE, results = "hide"}
(function(x) {x + 3}) (2)
```

Often anonymous functions are written without the `{}`, like this:
```{r echo = TRUE, results = "hide"}
(function(x) x + 3) (2)
```

<!--
Here's an anonymous function for calculating the mean of a vector `x`. In the following example, the input `x` to the function is each element of the list `l`.
```{r}
l <- list(1:5, 5:7)
lapply(l, FUN = function(x){sum(x)/length(x)})
```

see http://adv-r.had.co.nz/Functional-programming.html#anonymous-functions
-->

---

## Anonymous Functions with a vector as input

Same function as before but here we pass a vector as input rather than a single number:
```{r echo = TRUE, results = "hide"}
add <- function(x) {
  return(x + 3) }
add(c(1,2,3))
```

To rewrite it as anonymous function, the following code **won't work** because the vector isn't passed correctly:
```r
(function(x) x + 3) c(1, 2, 3)
```

Instead, **this works correctly**, since the entire vector is passed as a single argument:
```{r echo = TRUE, results = "hide"}
(function(x) x + 3) (c(1, 2, 3))
```

<!--
So the problem isn’t that anonymous functions can’t handle multiple values — they can!
The issue is the comma-separated syntax, which in R means multiple arguments, not a vector of values.
-->

---

## Anonymous Functions with sapply() and lapply()

You can also rewrite the previous code using `sapply()` or `lapply()`. Remember them from last time? They can be used instead of for loops in R.

Both belong to the `apply()` family and have the same basic structure:

```r
lapply(X, FUN)
sapply(X, FUN)
```

- `X` is a vector or list (the input)
- `FUN` is a function to apply to each element of X

They both loop through the elements of `X`, apply `FUN` to each element...

---

## Anonymous Functions with sapply() and lapply()

...But then differ in how they return the result:

- `lapply`  means "list apply", always returns the output as a list
- `sapply()` means "simplify apply", returns a simplified version of the list (usually a vector, sometimes a matrix)

```{r echo = TRUE, results = "hide"}
# previous anonymous function
(function(x) x + 1) (c(1, 2, 3))

# re-written with sapply
sapply(c(1,2,3), function(x) x + 1)

# re-written with lapply
lapply(c(1,2,3), function(x) x + 1)
```

**Tip:** Using `sapply()` is more common because it returns a vector, which is easier to work with that a list!

---

## Anonymous Functions the purrr way

The `purrr` package (remember the `map()`, `map_dbl()` from last lecture? same package!) provides another way to write anonymous functions.

Same function as before that adds 1 to each element of the input:
```{r echo = TRUE, results = "hide"}
add <- function(x) {
  x + 1
}
add(c(1,2,3))
```

Rewrite it as anonymous function in [`purrr`](https://purrr.tidyverse.org/reference/map.html) in two ways:
```{r echo = TRUE, results = "hide"}
library(purrr)
map_dbl(c(1, 2, 3), function(x) x + 1)
map_dbl(c(1,2,3), ~.x + 1)
```

*Tip:* `map()` has several versions like `map_dbl()`, `map_chr()`, and `map_df()` which let you control the output type directly

---

## 💻 Practice writing functions 

**Download the class materials from the website (see Lecture 18 homepage): complete QUESTION 3. Once done, share your code [here](https://codeshare.io/2KqD88)**



# 2. Documenting your function {.slide .center .middle}

---

## Functions should be well-documented! How? 

Each function should include the following:

* a one-sentence description of the function summarizing what it does
* all function's arguments, denoted as `Args:`, with their data types
* a description of the return value, denoted by `Returns:` 

**Do not use AI do generate your documentation: it overly complicate things** 

---

## Codying style: document your function

```{r echo = TRUE, results = "hide"}

my_sum <- function(vector) {
  
  # Calculates the sum of all numbers in a numeric vector using a `for` loop.
  # Args:
  #   vector (num): a numeric vector whose elements will be summed
  # Returns:
  #  value (num): single numeric value.
  # Example use:
  #   x <- c(3, 1, 4)
  #   my_sum(x)
    
  total <- 0
  for (i in seq_along(vector)) {
    total <- total + vector[i]
  }
  total
}
```

---

## Example from Google's R Style Guide [here](https://web.stanford.edu/class/cs109l/unrestricted/resources/google-style.html#:~:text=Function%20Documentation,-Functions%20should%20contain&text=These%20comments%20should%20consist%20of,value%2C%20denoted%20by%20Returns%3A%20.)

```{r echo = TRUE, results = "hide"}

CalculateSampleCovariance <- function(x, y, verbose = TRUE) {

  # Computes the sample covariance between two vectors.
  # Args:
  #   x: One of two vectors whose sample covariance is to be calculated.
  #   y: The other vector. x and y must have the same length, greater than one,
  #      with no missing values.
  #   verbose: If TRUE, prints sample covariance; if not, not. Default is TRUE.
  # Returns:
  #   The sample covariance between x and y.
  
  n <- length(x)
  if (n <= 1 || n != length(y)) {
    stop("Arguments x and y have invalid lengths: ",
         length(x), " and ", length(y), ".") }
  if (TRUE %in% is.na(x) || TRUE %in% is.na(y)) {
    stop(" Arguments x and y must not have missing values.") }
  covariance <- var(x, y)
  if (verbose)
    cat("Covariance = ", round(covariance, 4), ".\n", sep = "")
  return(covariance)
}
```

---

## 💻 Practice writing functions 

**Download the class materials from the website (see Lecture 18 homepage)**

**AFTER CLASS CHALLENGE: complete QUESTION 5 and QUESTION 6 on writing a more complex function and adding documentation to it** 



# 3. Writing functions with the tidyverse {.slide .center .middle}

---

## Functions and {{ }}

You can write custom functions and use them together with `dplyr` or `ggplot` functions!

You need the following: 

* syntax to write a custom function
* special operators that allows R to evaluate your custom functions correctly: 
   * bang-bang `!!`  with `enquo()` (this is the old method)
   * curly-curly or embracing `{{ }}` (this replaced the former method!)

Check out Chapter 25 of our book for details <https://r4ds.hadley.nz/functions.html>

---

## Functions and {{ }}

Example: create a data frame of age and height of 5 individuals:
```{r echo = TRUE}
library(dplyr)

df <- tibble(age = c(10, 12, 13, 14, 10), 
             height = c(110, 140, 155, 170, 130))
df
```

---

## Functions and {{ }}

Use `summarize()` from `dplyr` to compute the mean and standard deviation on the column `age` of this data frame:

```{r echo = TRUE}

df %>%
  summarize(mean_age = mean(age),
            sd_age = sd(age))
```

---

## Functions and {{ }}

Rewrite the previous code using a function and the curly-curly operator (embracing): 

```{r echo = TRUE}

custom_summary <- function(data, col) {
  data %>%
    summarize(mean_value = mean({{ col }}),
              sd_value = sd({{ col }}))
}

custom_summary(df, age)
```

The `{{ }}` operator tells `dplyr` not to treat "col" as the literal variable name, but instead to look inside it for the variable we actually need (here "age")

---

## Functions and {{ }}

**Embracing a variable means to wrap it in braces so `var` becomes `{{ var }}` when using functions `dplyr` or `ggplot` code.**

It tells `dplyr` to use the value stored inside the argument, not the argument as the literal variable name

From the book: "one way to remember what’s happening is to think of `{{ }}` as looking down a tunnel — `{{ var }}` will make a `dplyr` function look inside of var rather than looking for a variable called var"

---

## Functions and {{ }}

Rewrite it with multiple columns as arguments:

```{r echo = TRUE}

custom_summary <- function(data, col1, col2) {
  results <- list(
    summarise(data, 
              mean_value = mean({{ col1 }}),
              sd_value = sd({{ col1 }})),
    summarise(data, 
              mean_value = mean({{ col2 }}),
              sd_value = sd({{ col2 }}))
  )
  # Combine output into a single data frame
  bind_rows(results)
}
custom_summary(df, age, height)
```

---

## Functions and {{ }}

**When do you need to embrace when writing a function?**

In all cases of "tidy evaluation," which has two main types (see the readings, Chapter 25 for details): 

* Data Masking: in functions such as `arrange()`, `filter()`, `summarize()` from `dplyr` and in `aes()` from `ggplot2` often *when you need to pass a variable name*

* Tidy Selection: in functions such as `select()`, `relocate()`, `rename()` *when you need to select variables by name* 

<!--
### Functions and `{{ }}`: additional resources

**Using bang-bang `!!`**

* Chapter ["7 Tidy evaluation"](https://krlmlr.github.io/tidyprog/tidy-evaluation.html): explanation with examples on how and why to use `!!` with `enquo`

* Blog post ["Writing Custom Tidyverse Functions"](https://jonthegeek.com/2018/06/04/writing-custom-tidyverse-functions/): explanation with step-by-step example on how and when to use `!!` with `ensym` and with `enquo`

* Another example [here](http://zevross.com/blog/2018/09/11/writing-efficient-and-streamlined-r-code-with-help-from-the-new-rlang-package/) 

**Using curly-curly `{{ }}`**

* Blog post ["Curly-Curly, the successor of Bang-Bang"](https://www.r-bloggers.com/2019/06/curly-curly-the-successor-of-bang-bang/)

* Another example [here](https://agstats.io/post/writing-r-functions/) (under "Functions and Tidy Evaluation")
-->

---

## 💻 Practice writing functions 

**Download the class materials from the website (see Lecture 18 homepage): complete QUESTION 7. Once done share your code [here](https://codeshare.io/anVgz4)**

---

## To print these slides as pdf

Click on the icon bottom-left corner (the three lines) \> Tools \> PDF Export Mode \> Print as a Pdf

