---
title: "Computing for the Social Sciences: Lecture 20"
subtitle: "Topics: Reproducible Workflow"
author: "Sabrina Nardin, Summer 2026"
format:
  revealjs:
    theme: simple
    slide-number: true
    incremental: false
    css: ../style/styles.css
    code-overflow: wrap
    highlight-style: github
    ratio: 16:9
    chalkboard: true
    toc: false
    center: true
---

# Agenda {.center}

::: {.agenda-list}
1. Tips to improve your R Workflow
2. Deep-dive into R Markdown
3. Git/Github with the command line
:::

<span style="font-size: 0.8em; color: #666;">
*Slides last updated on **`r format(Sys.Date(), '%B %d, %Y')`**. Slides authored by Sabrina Nardin. AI used to polish slides style and fix typos.*
</span>

<!--

## Reflecting on AI Use for Programming

One part of my job is teaching you R fundamentals. Another (new) part is helping you learn how to use AI as a learning tool, as I think AI is here to stay.

I'd like to do an activity on reflecting on your AI use in this course using this [Google Doc](https://docs.google.com/document/d/1RyzfQ022OdChYVKWW6VPhEnJgjUmdyYmhtumHld-cso/edit?usp=sharing)


### Slides are in progress and will be updated before class.

::: {.content-hidden}

# 1. Exam 2 Overview + Final Project Check-in {.slide .center .middle}

---

## Exam Format

### Identical to Exam 1 or very similar. Multiple-choice questions and short open-ended questions, such as:

- *Classical multiple choice* — select the correct answer from several options  
- *Spot the typo* — identify and fix syntax or logic errors
- *Matching questions* — pair code snippets with their descriptions or outputs  
- *Debugging questions* — code runs but gives unexpected results; explain *why*  
- *Predict the output* — given a short R snippet determine what the code will return
- *Open-ended questions or write code questions* — brief written explanations and interpretations of results; write code (possible topics: conditionals, loops, functions, base R, data structures)  

---

## Exam Logistics

- Topics covered: Lectures 9-17 (inclusive)

- Time: 60 minutes

- It is closed-book, but you may bring pen and paper for scratch work. See Syllabus for details

- Arrive on time: If you arrive late, you can still take the exam, but no additional time can be given

- Make-up and absence policy: See Syllabus

---

## How to Prepare for the Exam

1. **Review class materials:** slides, in-class code, and assigned readings  

2. **Review previous homework (especially 3 and 4)** re-run your code and check you understand each step  

3. **Review the in-class prep quizzes we have done** exam questions will be similar 

4. **Use AI strategically:** once you’ve done all the above, ask AI to generate additional practice questions of different types and level of difficulty. Starting with AI before reviewing materials can backfires

::: {.callout-tip}
Preparation works best when you write and run code yourself, review your mistakes, and make sure you can explain *why* each answer is correct.
:::

---

## Final Project Check-In

### When: 
Wednesday, December 3, 2025 after the exam (~2:30–3:20 PM)

### What: 
Optional 1:1 meetings (~10 minutes) with the instructor or TA to discuss your final project. Please come with whatever you’ve completed so far and any questions you’d like to go over.

### Sign up:
First come, first served. If many students plan to attend, a sign-up link will be provided.

-->


# 1. Tips to improve your R workflow {.slide .center .middle}

---

## Main Idea: Reproducibilty

Reproducibility: **you, or anyone else, can re-run your analysis from scratch and get the same result.**

The tips in the next slides all serve one principle:

::: {.callout-important}
**Don't preserve the workspace. Preserve the code that creates the workspace.**

Your scripts are the product. Everything else (objects in memory, the console history, the current session) is disposable. See Bryan & Hester, [*What They Forgot to Teach You About R*](https://rstats.wtf/)
:::

---

## Tip 1: Start every session with a clean, empty workspace

When you run an R session, your **workspace** includes everything created during that session:

- packages you loaded with `library()`
- all user-created objects (data, vectors, functions, models, ...)

By default, R offers to *save* this workspace when you quit and *restore* it when you return. This is convenient in the short run, but not in the long run.  

---

## Tip 1: How?

When closing a session, R asks: *"Do you want to save your workspace?"* → **Answer No.**

Even better, change the defaults so R never asks again: **Tools → Global Options → General**

- Uncheck "Restore .RData into workspace at startup"
- Set "Save workspace to .RData on exit" to **Never**

<!--
::: {.callout-tip}
You can also run `usethis::use_blank_slate()` to apply these settings for you.
:::
-->

---

## Tip 1: Why?

A restored workspace can hide objects you forgot about or that your current script never creates:

- objects created by *other* scripts may still be lurking
- packages or data may be loaded that your script never loads
- you may have modified objects directly in the Console without updating the script

If your script only "works" because of leftovers in the workspace, it is silently broken. That hurts reproducibility

<!-- When everything lives in the code, you, and anyone else, can rebuild the entire workspace from scratch, on any machine, at any time. -->

---

## Tip 2: Use a project-based workflow and relative paths

In this class, we use a project-based workflow **by creating an `.Rproj` file in every repo!** 
Value of this approach:

- keeps all files for one project together in one folder
- sets the working directory automatically, so **relative paths just work**
- lets you switch between projects (or open several) without confusion

::: {.callout-note}
If you complete HW5, remember to create `.Rproj` file in your repo: see [these slides from Lecture 10](https://brinasab.github.io/css-website-summer/lectures/10-slides.html#/rstudio-projects-make-using-relative-file-paths-easier).
:::

---

## Tip 2: Use a project-based workflow and relative paths

**Avoid this pattern:**

```r
# fragile: works only on this computer, only today
setwd("/Users/sabrina/Documents/css/hw5/data")
df <- read_csv("survey.csv")
```

An absolute path or a path hard-wired like this example will not work on anyone else's computer, or on yours a year from now.

Rely on the project root using the `.Rproj` folders. For more tips, explore the [`here` package](https://here.r-lib.org/) builds paths from the top-level project folder, no matter which script or subfolder you are in.

<!--
```r
# robust: relative to the .Rproj folder, works anywhere
library(here)
df <- read_csv(here("data", "survey.csv"))
```
-->


---

## Tip 3: Treat raw data as read-only

Your raw data files are your starting point: never edit them by hand, and never let a script overwrite them.

- keep raw files in their own folder (e.g., `data/raw/`) and treat it as untouchable
- do all cleaning **in a script**, so every change is recorded and repeatable
- save cleaned/processed versions to a separate folder (e.g., `data/processed/`)

::: {.callout-tip}
## Test
If you deleted everything in `data/processed/`, could you regenerate it by re-running your scripts on the raw data? If yes, your pipeline is reproducible. See Wilson et al. (2017), [*Good Enough Practices in Scientific Computing*](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1005510)
:::

---

## Tip 4: Build debugging skills

**What debugging tips have you learned from working with R so far?**

---

## Tip 4: Build debugging skills

**1. Realize that you have a bug:** 

- a bug can be syntactic (like a typo) 
- or conceptual (a misunderstanding of R's rules)

<!--
We all encounter both types of bugs. Fixing the latter requires time and experience, but we can address syntactic bugs faster by improving our coding style!
-->

---

## Tip 4: Build debugging skills

**2. Figure out where it is:**

- read your code out loud
- use `print()` statements to check each line
- run one piece of code at a time
- if it comes from misuse of a specific function, check its documentation with `?function_name`
- ask AI to identify it and explain why
- take a break!

**3. Fix it:** then re-run the *whole* script in a fresh session to confirm

More in ["Debugging R code"](https://rstats.wtf/debugging-r) from *What They Forgot to Teach You About R*

---

## Tip 5: Read and understand what R tells you

R communicates in three distinct ways, and they are **not** all bad news:

- **Errors** stop your code. R could not do what you asked: this must be fixed.

- **Warnings** let the code run, but flag that something may be off. Ceck the result before trusting it.

- **Messages** are informative. Nothing is wrong, R is just telling you what it did.

---

## Tip 5: Read and understand what R tells you

```r
log("a")
#> Error in log("a"): non-numeric argument to mathematical function
```
Errors: code stopped: fix it before moving on

```r
log(-1)
#> Warning message: NaNs produced
```
Warning: code ran, but the result is `NaN` (check your input)

<!--
log(-1) has no answer in the real numbers. But from R's perspective, the call itself is legitimate: log() expects a number, and you gave it a number. So R can carry out the computation — it just produces NaN ("Not a Number"), which is a real value in R that represents an undefined mathematical result. The code ran and returned something, so R warns you rather than stopping.
-->
```r
library(tidyverse)
#> ── Attaching core tidyverse packages ──────────────
```
Message: nothing is wrong, just useful info

<!--
For example, plotting with geom_smooth() prints a message because it automatically decides which smoothing algorithm to use (default depends on sample size):

```{r echo = TRUE, out.width = "35%"}
library(tidyverse)
ggplot(diamonds, aes(carat, price)) +
  geom_point() +
  geom_smooth()
```
-->

---

## Tip 6: Restart R often (Session → Restart R)

Stuck with a bug, but you're 100% sure the code is correct? Restart R:

- More reliable than `rm(list = ls())` or the broom icon. Those only delete user objects; packages you loaded, options you changed, and other session state all survive invisibly
- restarting keeps RStudio open but clears everything underneath, giving you a truly clean session
- if your script still works after a restart, run top to bottom, that's evidence it is reproducible

::: {.callout-note}
Restart early, restart often. If restarting feels costly, that is a sign your workflow depends on hidden state, exactly what Tips 1 and 2 prevent.
:::

---

## Tip 7: Use code auto-formatting in RStudio

Consistent style makes bugs easier to see (yours *and* others). RStudio helps:

- option 1: `Code > Reformat Code` (Shift + Cmd/Ctrl + A)
- option 2: `Code > Reindent Lines` (Cmd/Ctrl + I)
- option 3: install the [`styler`](http://styler.r-lib.org/) package for more on restyling

Open an R script and try this out:

```r
 y<-10
  if (y < 20) {
x <- "Too low" }
else {
  x <-"Too high"}
```

---

## Recap: habits for reproducible work

1. Start every session with a **clean, empty workspace**
2. Use a **project-based** workflow (`.Rproj` in every repo)
3. Treat raw data as **read-only**; clean it with scripts
4. Build **debugging** skills
5. Distinguish **errors, warnings, and messages**
6. **Restart R** often
7. **Auto-format** your code

::: {.callout-important}
**FINAL TEST: Close R, reopen the project, run your script top to bottom. Does it work?**
:::

---

## Resources for reproducible work

- Bryan & Hester, [*What They Forgot to Teach You About R*](https://rstats.wtf/) on clean workspaces, safe paths, debugging
- Bryan, ["Project-oriented workflow"](https://www.tidyverse.org/blog/2017/12/workflow-vs-script/) (tidyverse blog)
- Wickham, Çetinkaya-Rundel & Grolemund, [*R for Data Science*, "Workflow: scripts and projects"](https://r4ds.hadley.nz/workflow-scripts.html)
- Wilson et al. (2017), [*Good Enough Practices in Scientific Computing*](https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1005510), for general tips on the topic of reproducibility
- The [`here`](https://here.r-lib.org/) and [`styler`](http://styler.r-lib.org/) packages


# 2. Deep-dive into R Markdown {.slide .center .middle}

---

## R Markdown Recap

**R Markdown is a file format (with extension `.Rmd`) that:**

- combines R code, using code chunks, and text in a single document
- makes it easy to create reproducible reports, analyses, and presentations
- you might hear about Quarto (with extension `.qmd`): a newer alternative to R Markdown (if you learn R Markdown, you can quickly adjust to Quarto)

---

## R Markdown Recap

### Three Main Components

1.  **YAML header** for document title, author, output format, etc. Insert it at the top between `---`

2.  **Text** regular Markdown for headers, lists, links, etc.

3.  **Code chunks** for code. Insert it with triple backticks ```` ``` ````

::: {.fragment .setup-box .callout-tip title="Code Chunk Tips"}
**Three ways to insert a code chunk in your R Markdown:**

-   Use the RStudio toolbar\
-   Type manually ```` ```{r} ```` and ```` ``` ````
-   Use a shortcut
    -   Mac: ⌘ + ⌥ + I\
    -   Windows: Ctrl + Alt + I
:::

---

## R Markdown vs R Script

### When to use a script vs. a Rmd

- When you want to experiment, test functions, debug code

- When you do not need to communicate output or results to others

- For scripts that need to run top-to-bottom without narrative

- For larger projects where you could have some code in R scripts (e.g., data cleaning, analysis pipelines, tests, etc.), and others (e.g., the final-narrated versions) in a Rmd file

---

## R Markdown: What We Know

* Inserting a YAML Header with basic elements
* Formatting text (italics, bold, etc.)
* Formatting titles 
* Inserting lists (ordered and unordered)
* Inserting images, tables, and hyperlinks
* Rendering a Rmd into pdf, html, md, etc.

---

## R Markdown: What We Learn Today

* How the knitting process works
* Inline code
* Code chunks names
* Code chunks options
* Global options
* New YAML header elements

---

## R Markdown: The Knitting Process

When you "knit" your document, the following sequential things happen, as [illustarated here](https://bookdown.org/yihui/rmarkdown-cookbook/rmarkdown-process.html)

1.  R Markdown sends the `.Rmd` file to `knitr` http://yihui.name/knitr/

2.  Knitr executes all of the code chunks and creates a new plain Markdown `.md` file which includes the code and its output

3.  This plain Markdown file is then converted by `pandoc` into any number of output types including html, PDF, Word document, etc.: http://pandoc.org/

---

## R Markdown: Inline Code - How it looks like in your Rmd

````{verbatim, echo = TRUE}
---
title: "Gun deaths"
author: "Sabrina Nardin"
output: html_document
---

```{r setup, include = FALSE}
library(tidyverse)
library(rcis)
```

## Gun deaths by age

The `gun_deaths` dataset includes data for `r nrow(gun_deaths)` individuals killed by guns.

```{r youth, echo = FALSE}
youth <- gun_deaths %>% filter(age <= 65)
```

We have `r nrow(youth)` individuals who are 65 or younger. 
Only `r nrow(gun_deaths) - nrow(youth)` are older than 65.

````

---

## R Markdown: Inline Code - How is rendered once knitted

```{r youth, include = FALSE}
library(tidyverse)
library(rcis)
youth <- gun_deaths |> filter(age <= 65)
```

The `gun_deaths` dataset includes data for `r nrow(gun_deaths)` individuals killed by guns.

We have `r nrow(youth)` individuals who are 65 or younger. 
Only `r nrow(gun_deaths) - nrow(youth)` are older than 65.

---

## R Markdown: Code Chunks Names

A named code chunk:
````{verbatim, lang = 'markdown', echo = TRUE}
```{r youth-distribution}
# code goes here
```
````

Advantages:

1. Easier to navigate to specific chunks (using the drop-down code navigator in the script editor). Note: Two code chunks should not have the same name.

1. Graphics produced by the chunks have useful names

1. Use it with `cache = TRUE` to avoid re-performing expensive computations on every run 

::: {.callout-tip}
`cache = TRUE` saves a chunk's output to a file on memory. On future renders, R checks if the code chunk has changed; if not, R skips re-running the chunk and reuses the cached results (see Ch. 27 R for Data Science)
:::

---

## R Markdown: Code Chunk Options {.center}

::: small
| Option | What it does | Default |
|------------------------|------------------------|------------------------|
| `eval = FALSE` | Code is not run, only shown; results do not appear. Useful for code examples or showing code with errors. | `eval = TRUE` |
| `include = FALSE` | Code is run, but both code and results are hidden. Useful for setup code you don't want to display. | `include = TRUE` |
| `echo = FALSE` | Code is run, results are shown, but code is hidden. Good when readers only need the output. | `echo = TRUE` |
| `error = TRUE` | Code is run and shown, even if it throws an error. Useful when demonstrating broken code. | `error = FALSE` |
| `message = FALSE` / `warning = FALSE` | Code is run, but messages and warnings are hidden from the output. | `message = TRUE` / `warning = TRUE` |
:::

Check [this table](https://r4ds.had.co.nz/r-markdown.html#chunk-options) and [options](http://yihui.name/knitr/options/) for more.

---

## R Markdown: Global Options

Use global chunk options when you want the same settings to apply to ALL code chunks in your R Markdown file.

Global options are defined in a code chunk at the top of your Rmd file with `knitr::opts_chunk$set()` 

For example, if you want to hide the code by default in your entire Rmd:

```r
knitr::opts_chunk$set(
  echo = FALSE    # shows results but hide codes 
)
```

---

## R Markdown: New YAML Header Elements

````{verbatim, echo = TRUE}
---
title: "Gun deaths"
author: "Sabrina Nardin"
output: html_document
---
````

---

## YAML Header: Table of Contents

````{verbatim, echo = TRUE}
---
title: "Gun deaths"
author: "Sabrina Nardin"
output: 
  html_document
    toc: true
    toc_depth: 2
---
````

---

## YAML Header: Apparence and Style

````{verbatim, echo = TRUE}
---
title: "Gun deaths"
author: "Sabrina Nardin"
output: 
  html_document
    theme: readable
    fig_width: 8
    fig_height: 6
---
````

---

## YAML Header: Date as string

We can add a date as a simple string, and manually modify it every time we modify our file:

````{verbatim, echo = TRUE}
---
title: "Gun deaths"
output: html_document
date: "December 1, 2025"
---
````

But we can also add it automatically!

---

## YAML Header: Date with lubridate 

````{verbatim, echo = TRUE}
---
title: "Gun deaths"
output: html_document
date: "`r lubridate::today()`"
---
````

For more on the `lubridate` package, see Chapter 16 of R for Data Science and [Lecture 12 slides](https://brinasab.github.io/css-website-summer/lectures/12-home.html)

---

## YAML Header: Date with Sys.Date

````{verbatim, echo = TRUE}
---
title: "Gun deaths"
output: html_document
date: "`r Sys.Date()`"
---
````

`Sys.Date` returns the current day in the current time zone. The default uses the format year/month/day.

---

## YAML Header: Date with Sys.time

````{verbatim, echo = TRUE}
---
title: "Gun deaths"
output: html_document
date: "`r format (Sys.time(), '%d, %B, %Y')`"
---
````

`Sys.time` returns an absolute date-time value which can be converted to various time zones, and it more easily is customizable. This format says we want the date as: numeric day, full name of the month, and full (4-digit) year, but it can be changed

---

## 💻 Practice with Rmd

Download today's class materials and complete the tutorial!

---

## R Markdown More Info and Tips

If you do the final project (HW5), try to experiment with a couple of the R Markdown options we have seen today!

For more, see also [R for Data Science 1st Edition](http://r4ds.had.co.nz):

- Chapter 27 "R Markdown"
- Chapter 28 "Graphics for communication"
- Chapter 29 "R Markdown formats" 


# 3. Git/Github from the command line {.slide .center .middle}

Download the **Git/Github from the command line tutorial from [HERE](https://drive.google.com/file/d/1cY48ez1rirI4qDhNccWqzisRfAigi4fB/view?usp=sharing)** to follow along or try it out after class! 

---

## To print these slides as pdf

Click on the icon bottom-left corner (the three lines) \> Tools \> PDF Export Mode \> Print as a Pdf
