---
title: "HW02: Mass Shootings in America"
draft: false
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

# Overview
*
**Due: ~~Thursday, October 23~~ Friday, October 24 by 11:59 PM (Chicago Time)**

In the first assignment you have demonstrated knowledge of your software setup, Git/GitHub via RStudio, and R Markdown. The goal of this second assignment is to practice transforming and exploring data with `dplyr` and `ggplot2`.


# Accessing and Cloning Your `hw02` Repository

::: {.callout-important}
This will work only if you have submitted your GitHub username (see Lecture 1) and accepted the email invitation to join our GitHub organization (check the email linked to your GitHub account). Make sure you’ve completed the setup under [Software Setup](../setup/index.qmd) and confirmed everything is working properly.
:::

**Accessing the repo:** 

1. Go at [this link](https://classroom.github.com/a/pky4Murt) to accept the invitation and create your private `hw02` repository on GitHub. Your repository will be named `hw02-<USERNAME>` and will be ready within a few seconds. If this does not work, see the note in red above.

2. Once the repository has been created, click the link provided to access it.


**Cloning the repo into your Workbench:**

1. Log into *Workbench* and start a new project:  
   *File > New Project > Version Control > Git*

2. In the *Repository URL* field:
   - Paste the URL of your GitHub repository. To find the URL, go to your `hw02` repo on GitHub, click the green Code button, and copy the SSH or HTTPS link (use the method you configured during setup: Workbench users must select SSH).
   - The project name should auto-fill. If not, type `hw02-<USERNAME>` manually.

3. Choose where to save the project directory locally. *Tip:* Avoid moving folders after cloning them.  

4. Check the box *"Open in new session"* (recommended for keeping projects separate and organized)

5. Click *Create Project*. This will:
   - Create a new folder on your machine  
   - Initialize a Git repository linked to GitHub  
   - Open an RStudio Project in a new session


# General Homework Workflow

See [Homework 1](hw01.qmd) for details. The workflow is the same for all assignments (e.g., accept the repo, edit the files, save, then Pull – Stage – Commit – Push).


# Assignment Description

### The mass shooting dataset

The United States experiences far more mass shooting events than any other developed country in the world. Policymakers, politicians, the media, activists, and the general public acknowledge the widespread prevalence of these events. However, effective policies to prevent such incidents should be grounded in empirical data.

In July 2012, in the aftermath of a mass shooting in a movie theater in Aurora, Colorado,
[Mother Jones](https://www.motherjones.com/politics/2012/07/mass-shootings-map/) published a report on mass shootings in the United States since 1982. Importantly, they provided the underlying data set as [an open-source database](https://www.motherjones.com/politics/2012/12/mass-shootings-mother-jones-full-data/) for anyone interested in studying and understanding the topic.

This dataset is already installed on Workbench.

<!--
This dataset in included in the [`rcis`](https://github.com/css-materials/rcis) library on GitHub. 

* If you are working on Workbench, you should have everything already installed. Simply load the library by typing in your console `library(rcis)`, then load the dataset by typing `data("mass_shootings")`. Type `?mass_shootings` for detailed information on the variables and other coding information. I'd suggest to work with R version 4.2 on Workbench

* If you are using R on your local computer, you first need to install the `rcis` by typing in your console `remotes::install_github("css-materials/rcis")`. If you don't already have the `remotes` library installed, you will get an error. Go back and install it first using `install.packages()`, then install `rcis`. Finally, the mass shootings dataset can be loaded using `data("mass_shootings")`. Use the help function in R `?mass_shootings` for detailed information on the variables and other coding information.
-->

### Answer the questions

Your repository includes specific and open-ended questions. Answer all of them.

::: {.callout-note}

- The questions—especially the open-ended ones—are designed to help you think like a programmer and data analyst. They won’t walk you through every step but encourage you to apply what you’ve learned and work independently. Translating what a question asks into code by breaking it down in smaller pieces is one important learning goal of this exercise

- This homework is designed to be completed over multiple sessions. Start early, and remember to save, commit, and push your work regularly.

- Do not use AI tools to generate R code from scratch or think for you. You may use it to debug (after trying on your own) or look up how functions work (check course materials first). Only submit code you fully understand. Reach out if you have questions, we are here to help!
:::


# Formatting Guide

### Formatting graphs

Your final graphs should be appropriate for sharing with outsiders. That means your graphs should have:

* [A title](http://r4ds.had.co.nz/graphics-for-communication.html#label)
* Labels on the axes (type `?labs` in your Console for details)
* Consider adopting your own color scales, [taking control of your legends (if any)](http://www.cookbook-r.com/Graphs/Legends_(ggplot2)/), playing around with [themes](https://ggplot2.tidyverse.org/reference/index.html#section-themes), and generally customizing your graphs to improve their visual appeal and clarity.


### Formatting tables

When presenting tabular data, consider using the `kable()` function from the `knitr` package to format the table for the final document. Here is how you use this function.

The code below displays a basic table summarizing where gun deaths occurred:
```{r packages, message = FALSE}
library(tidyverse)
library(knitr)
library(rcis)
```

```{r table-bad, echo = TRUE}
# calculate total gun deaths by location
count(mass_shootings, location_type)
```

This is the same code but we added `kable()` to format the table, add a caption, and label the columns. Observe what it does:
```{r table-good, format = "asis", echo = TRUE}
count(mass_shootings, location_type) %>%
  kable(
    caption = "Mass shootings in the United States by location",
    col.names = c("Location", "Number of incidents")
  )
```

Run `?kable` in your Console for additional options

<!-- We expect you to use this function for formatting tabular data in this and future assigments. -->


# Submit the Assignment

To submit your assignment on Canvas, follow these steps:

1. Before the deadline, push the final version of your work to your GitHub repository. *Tip:* don’t wait until the end to make your first commit—commit; stage, and push your work regularly as you go.

   Your GitHub repository should include (make sure to stage-commit-push all these files):
   - `mass-shootings.Rmd`: main file you will work on
   - `mass-shootings.md`: you will generate this file from the `.Rmd` by simply knitting it, like you did for HW01. The output format is already set to `github_document` keep it that way unless you encounter problems, in which case you can switch to `md_document` (this is rarely needed)
   - `mass-shootings_files/`: this folder contains all the graphs that you generated in your `.Rmd`

2. When you're ready to submit:
   - Copy the URL of your GitHub repository. It will look something like: `https://github.com/css-fall25/hw02-yourusername`
   - Submit that URL on Canvas under HW02. Do not upload any files to Canvas. We only need the repository link.


# Assessment

All homework assignments are evaluated using [this rubric](https://docs.google.com/spreadsheets/d/1RAgv09Wn62X7-xsSb5bGBmEqAMFKh_om/edit?usp=sharing&ouid=112534119211880791899&rtpof=true&sd=true). For more, see the *Assessment* section of the Syllabus. 

Below are further guidelines for this specific homework to help you assess your work before submitting it. In the past, **"Excellent"** or **"Very Good"** work included submissions that:

- Completed all parts of the assignment correctly and accurately.
- Included code that:
  - Ran without errors
  - Followed good style (clean, readable, well-organized)
  - Was well-documented but not over-commented
  - Used appropriate complexity that matched the prompt and course content
  - Demonstrated understanding of key concepts and functions
  - Demonstrated use of class resources 
  - Followed the course AI policies
- Included graphs and tables that were:
  - Well-chosen for the variable types
  - Clearly labeled, visually polished, and customized beyond defaults
  - Interpreted in a clear analysis
  - Critically interpreted, using the data to back up the points made
- Showed strong use of:
  - Required packages, showing curiosity to use them beyond the basics
  - R Markdown syntax (e.g., no rendering errors, appropriate use of syntax, etc.)
- Followed good GitHub practices:
  - Multiple informative commits showing progress over time
  - All required files present and displaying correctly, including rendered graphs

<!--

- "Excellent" under the "Achievement" element of the rubric might include correctly completing at least some of the optional parts of the questions.

*Needs improvement:* Doesn't complete all components. Code is too minimal, poorly written, and/or not documented. Plots are not correct and/or too minimal (e.g., same type of plot for each graph, or doesn't use plots appropriate for the variables being analyzed, or missing labels/titles, or leave default without experimenting). Does not follow or follows incorrectly the provided formatting guide. Shows partial or incomplete understanding of the packages needed for the assignment. Interpretation is lacking or incomplete. No record of commits other than the final push to GitHub (we expect students to commit often).

*Good:* Solid effort. Hits all the elements. Minor omissions but no clear mistakes. Easy to follow (both the code and the output). Shows sufficient understanding of the packages needed for the assignment.

*Excellent:* Finished all components of the assignment correctly. Code is well-documented (both self-documented and with additional comments as necessary). Graphs and tables are properly labeled, well-executed, and carefully chosen. Analysis is clear and easy to follow (e.g., graphs are labeled clearly and/or additional text to describe how you interpret the output). Shows solid understanding of the packages needed for the assignment. Repo shows history of multiple commits to back up and show a progression in the work.
-->

<!--
# Acknowledgments

The initial version of this homework was developed by Benjamin Soltoff  (“Computing for the Social Sciences” licensed under the CC BY-NC 4.0 Creative Commons License). Further implementations have been developed by Sabrina Nardin.

{r child = here::here("R", "_ack_ben.Rmd")}

[^clean]: For the purposes of this assignment, some data cleaning of the underlying raw data have been performed. You can view the data cleaning code [here](https://github.com/cis-ds/rcis/blob/master/data-raw/mass-shootings.R).
-->